### 2023-09-21 project-plan.Rmd
Comment on project plan

---
title: "Performance of prediction models for opportunities for improvment in trauma care among women compared to men and across patient cohorts"
author: "Anna Thoss"
bibliography: bibliography.bib
csl: vancouver.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "out") })
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- The project plan should be 3-4 pages long and written in
Rmarkdown (like this document). Remove all comments in italic when you 
use this document as a template for your own project plan. -->

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

# Introduction

Traumatic injuries<!-- Call it just trauma--> are a significant public
health concern worldwide, accounting for a substantial portion of
morbidity and mortality rates, being the cause of more than 5 million
fatalities each year [@WHO2014]<!-- This is an old reference, most
current figure is 4.3 million. Use:
@article{injuries2020,
  title =	 {Injuries—Level 1 cause},
  volume =	 {396},
  ISSN =	 {0140-6736},
  journal =	 {The Lancet},
  publisher =	 {Elsevier BV},
  author =	 {{GBD 2019 Diseases and Injuries Collaborators}},
  year =	 {2020},
  month =	 {Oct},
}
-->.  Trauma, and care of patients
with traumatic injuries, have long been a significant target point for
healthcare improvement programs [@WHO2009].  Multidisciplinary
mortality and morbidity conferences form the cornerstone of
initiatives dedicated to enhancing the quality of trauma care and,
consequently, improving patient outcomes [@Santana2014].

At mortality and morbidity conferences, representatives from different
professions and specialties adjacent to trauma care come together to
discuss care provided to a specific patient and compare it to care
given under optimal conditions. The findings of such a conference are
whether opportunities for improvement (OFI) can be found in any given
case.  Examples of opportunities for improvement<!-- You can use OFI
here, you abbreviated it in the previous sentence--> may include lack
of resources and management errors [@Lazzara2020].

Audit filters are applied to determine possible events in patient care
that are associated with unfavourable outcomes. The occurrence of such
an event triggers a review process, and when appropriate leads to the
implementation of corrective measures to address systematic errors and
feedback to individual practitioners [@McDermott_1994]. The use of
audit filters has its drawbacks, being associated with high rates of
false positives. [Saknar tydlig källa här]<!-- Consider changing the
order of these paragraphs, so that you introduce audit filters before
mortality and morbidity conferences, considering that audit filters
are used to select patients for those conferences-->

To enhance the precision of the selection process, there have been
efforts to introduce trauma mortality prediction models. Nevertheless,
the performance of these models has been unsatisfactory
[@Ghorbani2018] [@Heim_2016]<!-- When you cite more than one
reference, use [@Ghorbani2018; @Heim_2016]--><!-- Elaborate, how was
the performance unsatisfactory?--> . This is likely attributed to the
fact that<!-- "attributed to the fact that" can be just "because"-->
these models were originally developed with a focus on predicting
mortality rather than morbidity or instances of care failures within
the realm of trauma care. Machine learning-based prediction models
have been found to outperform conventional methods to predict
opportunities for improvement. *Hur gör jag med källhänvisning här?,
referera till preprint artikel?*<!-- You can do that-->

Men make up a disproportionally large group compared to women when it
comes to trauma. Each year, almost twice as many men die due to
traumatic injury when compared to women, with the distribution of
different types of injury varying by demographic and region
[@WHO2014].  It is important, that due to the differences between
these groups that the model for prediction has similar performance in
these subgroups.  Furthermore, it is important to establish if trained
models perform equally well across different cohorts within traumatic
injuries.<!-- Elaborate on these cohorts-->

# Aim

To assess the difference in performance of prediction models for
opportunities for improvement in trauma care among women compared to men
and across clinical trauma cohorts.

# Methods
<!-- I would find this section easier to read if you followed some
reporting guideline, like TRIPOD--> This is a registry-based study,
encompassing all trauma patients enlisted in both the Karolinska
University Hospital trauma registry and the trauma care quality
database spanning from 2012 to 2022. The Karolinska University
Hospital in Solna is equivalent to a level 1 trauma centre and manages
approximately 1,500 acute trauma patients each
year[@Ghorbani2018][Swetrau, rapport]. The trauma registry comprises
all patients admitted to the hospital with trauma team activation,
irrespective of their Injury Severity Score (ISS), as well as patients
with an ISS exceeding 9. Included in the registry is data on vital
signs, injuries, interventions, times, and patient demographics.

The trauma registry encompasses approximately 12,000 patients treated
from 2012 to 2022. The trauma care quality database in turn is a subset
of the trauma registry, including about 6000 patients selected for
review in the years 2013-2022.

The outcome of this study is opportunities for improvement (OFI), which
are identified through meetings of the multidisciplinary review board
during mortality and morbidity conferences. An OFI is a consensus
decision made during a conference and includes proposals for corrective
actions. These in turn are recorded in the trauma care quality database.
The trauma care quality database also includes data relevant to
mortality and morbidity conferences, such as audit filters. The
mortality and morbidity conferences at Karolinska University Hospital
involve professionals from various specialties within trauma care,
including surgery, neurosurgery, orthopedics, anesthesia, intensive
care, nursing, and radiology. Cases considered for a conference go
through multiple rounds of review. Mortality leads to direct inclusion.
Between 2013 and 2017, non-fatal poor outcomes were identified through
reviews conducted by a specialized trauma nurse to identify potential
OFI. Starting in 2017, the process was formalized, involving an initial
review by a specialized trauma nurse along with the application of audit
filters. All cases that were flagged either during the initial nurse
review or by the audit filters underwent a secondary review conducted by
two specialized nurses. If the second review identified potential areas
for improvement (OFI), the case was marked for discussion in a
multidisciplinary conference.

All patients reviewed for possible opportunities for improvement between
the years 2013 and 2021 are included in the study, with the exception of
patients under the age of 15 due to differing clinical pathways in this
subgroup.

The outcome of this study will be defined as "Yes, there is a
significant difference in models prediction of OFI in different cohorts"
or "No, there is no significant difference in models prediction of OFI
in different cohorts." Data for this will be extracted from the trauma
care quality database and run through different previously trained
machine learning models. *osäker på om outcome här är hållbart
formulerad*<!--Good enough for now-->

All variables from the trauma registry will be considered as potential
predictors in our analysis<!-- Note that the models are already
developed. So you should write about what have already been done,
i.e. these variables are already incorporated into the
models-->. These variables include information spanning the
pre-hospital, initial care, and subsequent in-hospital phases, which
include initial vital signs, timing, and types of procedures and
interventions, duration and level of care, injury details, mechanisms
of injury, injury types, and standard demographic information. This
array of predictors comprises both continuous and categorical
variables, ultimately resulting in the final models utilizing a total
of 45 predictors. (infoga bilaga) All available data will be used,
with a sample size of 6310 patients. This is due to previously
developed prediction models being based on all available data, as well
as including all 45 predictors regardless of the learner used for the
development of the model.

The models have been developed and trained on the data set previously
to this study<!-- Add preprint citation-->. Eight models are included,
all being developed using the Tidymodels framework (). These include
logistic regression (LR), random forest (RF), decision tree (DT),
support vector machine with a radial basis kernel (SVM), XGBoost,
LightGBM, CatBoost, and k-nearest neighbor (k-NN).

The different models' performance were assessed in the following cohorts:

1.  Blunt multi-system trauma: Blunt trauma with AIS ≥ 3 in at least two
    of the following AIS body regions: head, face, neck, thorax,
    abdomen, spine, or upper and lower extremities.

2.  Isolated severe TBI: Injury isolated to the area of the brain with
    AIS \> 2 and:

    a.  pre-hospital GCS of \< 9

        or

    b.  Pre- or in-hospital intubation.

3.  Blunt multi-system trauma without TBI: Blunt trauma with AIS \> 2
    and injuries in at least two of the following AIS body regions:
    head, face, neck, thorax, abdomen, spine, or upper and lower
    extremities and no injury to the brain with AIS \>2 and:

    a.  pre-hospital GCS of \< 9

        or

    b.  Pre- or in-hospital intubation.

4.  Blunt multi-system trauma with TBI: Blunt trauma with AIS \> 2 and
    injuries in at least two of the following AIS body regions: head,
    face, neck, thorax, abdomen, spine, or upper and lower extremities
    and with injury to the brain with AIS \>2 and:

    a.  pre-hospital GCS of \< 9

        or

    b.  Pre- or in-hospital intubation

Along with these cohorts, all machine learning models will be assessed
in regard to differences between male and female sex in their
prediction of OFI.

We will employ a 95% confidence level and a 5% significance level for
our analysis, with all statistical analysis conducted using R.

# Project Update and Timeline

![**The Gantt chart will be displayed here when you place it in the
assets folder under "project-gantt.png"**](assets/project-gantt.png)

Gantt chart over the project timeline.

# Back-up Plan

Currently, there is no need to develop a thorough backup plan since all
the necessary data for the statistical analysis has been collected and
is accessible.

# References
=======

Introduction
============

*This is part A) in the project plan [sv: projektredogörelse]. *

*The background/introduction should summarise the scope of the problem,
what is known about the problem, what is not known, what specific
knowledge gap the study is supposed to fill and why this is
important. It should end be stating the aim of the study. The
background should be 4-5 paragraphs long and each paragraph should be
between 3-5 sentences. Strive to make the paragraphs about the same
length.*

*bibliography.bib includes an example reference. Add additional
references in that file as bibtex entries and then cite in text using:
"[@exampleKey9999]". One can convert DOIs' to bibtex entries using 
https://www.doi2bib.org/. If you use Zotero you can simply export 
entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*

Methods
=======

*Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then
use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*

Project Update and Timeline
===========================

*This is part B) in the project plan [sv: projektredogörelse]. *

*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*

![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)

Back-up Plan
============

*This is part C) in the project plan [sv: projektredogörelse]*

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2023-09-21 manuscript.Rmd
utkast project plan

---
title: "Manuscript title here"
author: "Your name here"
output:
  pdf_document: default
  html_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->

<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->

Abstract
========

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->

You can cite document here like this [@exampleKey9999]. Open the file
bibliography.bib to learn more.

Methods 
=======

Study design
------------

<!-- Present key elements of study design -->

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

Participants
------------

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

Bias 
----

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d)

Cohort study: If applicable, explain how loss to follow-up was addressed

Case-control study:If applicable, explain how matching of cases and
controls was addressed

Cross-sectional study: If applicable, describe analytical methods
taking account of sampling strategy

(e) Describe any sensitivity analyses
-->

Results
=======

<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Case-control study — Report numbers in each exposure category, or
summary measures of exposure

Cross-sectional study — Report numbers of outcome events or summary
measures

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

You can also mix text and code, so called inline code, like this: `r 2+5`.

Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2023-09-21 project-plan.Rmd
utkast project plan

@@
 title: "Performance of prediction models for opportunities for improvment in trauma care among women compared to men and across patient cohorts"
+subtitle: ""
 author: "Anna Thoss"
@@
 
+```{=html}
 <!-- The project plan should be 3-4 pages long and written in
@@
 use this document as a template for your own project plan. -->
-
+```
+```{=html}
 <!-- This is an R Markdown document. Markdown is a simple formatting
@@
 details on using R Markdown see http://rmarkdown.rstudio.com -->
-
+```
+```{=html}
 <!-- This is a comment and will not be present in the compiled
@@
 the console and you should only see headings -->
-
+```
 # Introduction
 
-Traumatic injuries<!-- Call it just trauma--> are a significant public
-health concern worldwide, accounting for a substantial portion of
-morbidity and mortality rates, being the cause of more than 5 million
-fatalities each year [@WHO2014]<!-- This is an old reference, most
-current figure is 4.3 million. Use:
-@article{injuries2020,
-  title =	 {Injuries—Level 1 cause},
-  volume =	 {396},
-  ISSN =	 {0140-6736},
-  journal =	 {The Lancet},
-  publisher =	 {Elsevier BV},
-  author =	 {{GBD 2019 Diseases and Injuries Collaborators}},
-  year =	 {2020},
-  month =	 {Oct},
-}
--->.  Trauma, and care of patients
-with traumatic injuries, have long been a significant target point for
-healthcare improvement programs [@WHO2009].  Multidisciplinary
-mortality and morbidity conferences form the cornerstone of
-initiatives dedicated to enhancing the quality of trauma care and,
-consequently, improving patient outcomes [@Santana2014].
+Traumatic injuries are a significant public health concern worldwide,
+accounting for a substantial portion of morbidity and mortality rates,
+being the cause of more than 5 million fatalities each year [@WHO2014].
+Trauma, and care of patients with traumatic injuries, have long been a
+significant target point for healthcare improvement programs [@WHO2009].
+Multidisciplinary mortality and morbidity conferences form the
+cornerstone of initiatives dedicated to enhancing the quality of trauma
+care and, consequently, improving patient outcomes [@Santana2014].
 
@@
 professions and specialties adjacent to trauma care come together to
-discuss care provided to a specific patient and compare it to care
-given under optimal conditions. The findings of such a conference are
-whether opportunities for improvement (OFI) can be found in any given
-case.  Examples of opportunities for improvement<!-- You can use OFI
-here, you abbreviated it in the previous sentence--> may include lack
-of resources and management errors [@Lazzara2020].
+discuss care provided to a specific case and compare it to care given
+under optimal conditions. The findings of such a conference are whether
+opportunities for improvement (OFI) can be found in any given case.
+Examples of opportunities for improvement may include lack of resources
+and management errors [@Lazzara2020].
 
@@
 implementation of corrective measures to address systematic errors and
-feedback to individual practitioners [@McDermott_1994]. The use of
-audit filters has its drawbacks, being associated with high rates of
-false positives. [Saknar tydlig källa här]<!-- Consider changing the
-order of these paragraphs, so that you introduce audit filters before
-mortality and morbidity conferences, considering that audit filters
-are used to select patients for those conferences-->
+feedback to individual practitioners [@McDermott_1994]. The use of audit
+filters has its drawbacks, being associated with high rates of false
+positives. [Saknar tydlig källa här]
 
@@
 efforts to introduce trauma mortality prediction models. Nevertheless,
-the performance of these models has been unsatisfactory
-[@Ghorbani2018] [@Heim_2016]<!-- When you cite more than one
-reference, use [@Ghorbani2018; @Heim_2016]--><!-- Elaborate, how was
-the performance unsatisfactory?--> . This is likely attributed to the
-fact that<!-- "attributed to the fact that" can be just "because"-->
-these models were originally developed with a focus on predicting
-mortality rather than morbidity or instances of care failures within
-the realm of trauma care. Machine learning-based prediction models
-have been found to outperform conventional methods to predict
-opportunities for improvement. *Hur gör jag med källhänvisning här?,
-referera till preprint artikel?*<!-- You can do that-->
+the performance of these models has been unsatisfactory [@Ghorbani2018]
+[@Heim_2016]. This is likely attributed to the fact that these models
+were originally developed with a focus on predicting mortality rather
+than morbidity or instances of care failures within the realm of trauma
+care. Machine learning-based prediction models have been found to
+outperform conventional methods to predict opportunities for
+improvement. *Hur gör jag med källhänvisning här?, referera till
+preprint artikel?*
 
@@
 traumatic injury when compared to women, with the distribution of
-different types of injury varying by demographic and region
-[@WHO2014].  It is important, that due to the differences between
-these groups that the model for prediction has similar performance in
-these subgroups.  Furthermore, it is important to establish if trained
-models perform equally well across different cohorts within traumatic
-injuries.<!-- Elaborate on these cohorts-->
+different types of injury varying by demographic and region [@WHO2014].
+It is important, that due to the differences between these groups that
+the model for prediction has similar performance in these subgroups.
+Furthermore, it is important to establish if trained models perform
+equally well across different cohorts within traumatic injuries.
 
@@
 # Methods
-<!-- I would find this section easier to read if you followed some
-reporting guideline, like TRIPOD--> This is a registry-based study,
-encompassing all trauma patients enlisted in both the Karolinska
-University Hospital trauma registry and the trauma care quality
-database spanning from 2012 to 2022. The Karolinska University
-Hospital in Solna is equivalent to a level 1 trauma centre and manages
-approximately 1,500 acute trauma patients each
-year[@Ghorbani2018][Swetrau, rapport]. The trauma registry comprises
-all patients admitted to the hospital with trauma team activation,
+
+This is a registry-based study, encompassing all trauma patients
+enlisted in both the Karolinska University Hospital trauma registry and
+the trauma care quality database spanning from 2012 to 2022. The
+Karolinska University Hospital in Solna is equivalent to a level 1
+trauma centre and manages approximately 1,500 acute trauma patients each
+year[@Ghorbani2018][Swetrau, rapport]. The trauma registry comprises all
+patients admitted to the hospital with trauma team activation,
 irrespective of their Injury Severity Score (ISS), as well as patients
@@
 machine learning models. *osäker på om outcome här är hållbart
-formulerad*<!--Good enough for now-->
+formulerad*
 
 All variables from the trauma registry will be considered as potential
-predictors in our analysis<!-- Note that the models are already
-developed. So you should write about what have already been done,
-i.e. these variables are already incorporated into the
-models-->. These variables include information spanning the
-pre-hospital, initial care, and subsequent in-hospital phases, which
+predictors in our analysis. These variables include information spanning
+the pre-hospital, initial care, and subsequent in-hospital phases, which
 include initial vital signs, timing, and types of procedures and
-interventions, duration and level of care, injury details, mechanisms
-of injury, injury types, and standard demographic information. This
-array of predictors comprises both continuous and categorical
-variables, ultimately resulting in the final models utilizing a total
-of 45 predictors. (infoga bilaga) All available data will be used,
-with a sample size of 6310 patients. This is due to previously
-developed prediction models being based on all available data, as well
-as including all 45 predictors regardless of the learner used for the
+interventions, duration and level of care, injury details, mechanisms of
+injury, injury types, and standard demographic information. This array
+of predictors comprises both continuous and categorical variables,
+ultimately resulting in the final models utilizing a total of 45
+predictors. (infoga bilaga) All available data will be used, with a
+sample size of 6310 patients. This is due to previously developed
+prediction models being based on all available data, as well as
+including all 45 predictors regardless of the learner used for the
 development of the model.
 
-The models have been developed and trained on the data set previously
-to this study<!-- Add preprint citation-->. Eight models are included,
-all being developed using the Tidymodels framework (). These include
-logistic regression (LR), random forest (RF), decision tree (DT),
-support vector machine with a radial basis kernel (SVM), XGBoost,
-LightGBM, CatBoost, and k-nearest neighbor (k-NN).
+The models have been developed and trained on the data set previously to
+this study. Eight models are included, all being developed using the
+Tidymodels framework (). These include logistic regression (LR), random
+forest (RF), decision tree (DT), support vector machine with a radial
+basis kernel (SVM), XGBoost, LightGBM, CatBoost, and k-nearest neighbor
+(k-NN).
 
-The different models' performance were assessed in the following cohorts:
+To assess different models' performance across different cohorts, data
+from different cohorts run through multiple models. These cohorts are
+defined as follows:
 
@@
 Along with these cohorts, all machine learning models will be assessed
-in regard to differences between male and female sex in their
+in regard to differences between male and female gender in their
 prediction of OFI.
@@
 # References
-=======
-
-Introduction
-============
-
-*This is part A) in the project plan [sv: projektredogörelse]. *
-
-*The background/introduction should summarise the scope of the problem,
-what is known about the problem, what is not known, what specific
-knowledge gap the study is supposed to fill and why this is
-important. It should end be stating the aim of the study. The
-background should be 4-5 paragraphs long and each paragraph should be
-between 3-5 sentences. Strive to make the paragraphs about the same
-length.*
-
-*bibliography.bib includes an example reference. Add additional
-references in that file as bibtex entries and then cite in text using:
-"[@exampleKey9999]". One can convert DOIs' to bibtex entries using 
-https://www.doi2bib.org/. If you use Zotero you can simply export 
-entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
-Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*
-
-Methods
-=======
-
-*Refer to the appropriate reporting guideline for details. If you are
-developing, updating or validating a clinical prediction model then
-use
-[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
-you are conducting an observational study, for example a cohort or
-case control study in which you assess associations between some
-exposure and an outcome then use
-[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*
-
-Project Update and Timeline
-===========================
-
-*This is part B) in the project plan [sv: projektredogörelse]. *
-
-*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*
-
-![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)
-
-Back-up Plan
-============
-
-*This is part C) in the project plan [sv: projektredogörelse]*
-
-References
-==========
 

### 2023-09-21 main.R

@@ -18,9 +18,13 @@
 ## them). They all need to be sourced however when you compile your
 ## manuscript file or run this file as a job, as that happens in a
 ## clean R session.
+library(rofi)
 noacsr::source_all_functions()
 
 ## Import data
 data <- import_data(test = TRUE)
+
+combined.datasets <- rofi::merge_data(data, test = TRUE)
+combined.datasets$sofi <- rofi::create_ofi(combined.datasets)
 
 ## Whatever you do next, maybe clean data?

### 2023-09-23 project-plan.Rmd
Comment on project plan

--- 2023-09-21 project-plan.Rmd
@@
 
-Traumatic injuries<!-- Call it just trauma--> are a significant public
+Traumatic injuries<!-- Call it just trauma--> <!-- KS: when removing the word injuries it can be good to define "trauma" to differ from psychological trauma,
+A good citation is:
+@book{gerdin_risk_2015,
+    location = {Stockholm},
+    title = {The risk of dying: predicting trauma mortality in urban Indian
+             hospitals},
+    isbn = {978-91-7676-157-1},
+    shorttitle = {The risk of dying},
+    publisher = {Karolinska Institutet},
+    author = {Gerdin, Martin},
+    date = {2015},
+    note = {{OCLC}: 942942790},
+}
+A good read if you have time ;), PDF URL: https://openarchive.ki.se/xmlui/bitstream/handle/10616/44832/Thesis_Martin_Gerdin.pdf?sequence=3&isAllowed=y
+--> 
+are a significant public
 health concern worldwide, accounting for a substantial portion of
@@
 healthcare improvement programs [@WHO2009].  Multidisciplinary
-mortality and morbidity conferences form the cornerstone of
+mortality and morbidity <!-- KS: The order "morbidity and mortality" is mainly used in the literature. I also received a comment on this in my final version. --> conferences form the cornerstone of
 initiatives dedicated to enhancing the quality of trauma care and,
@@
 are used to select patients for those conferences-->
+<!-- KS: you will probably need to explain exactly what audit fitlers are.
+That they are a system that flags abnormal clinical parameters as possible OFI.
+These flag cases are then reviewed by two nurses and then a conference -->
 
@@
 <!-- I would find this section easier to read if you followed some
-reporting guideline, like TRIPOD--> This is a registry-based study,
+reporting guideline, like TRIPOD--> <!-- KS: prefer "We conducted a registry[...]" -->This is a registry-based study,
 encompassing all trauma patients enlisted in both the Karolinska
@@
 
-All patients reviewed for possible opportunities for improvement between
+All patients reviewed for possible opportunities for improvement <!-- KS: use OFI consistently after you have abbreviated it --> between
 the years 2013 and 2021 are included in the study, with the exception of
@@
 
+<!-- KS: Gianluigi will most likely not like this explaination so if you have time just come up with a random backup project -->
 Currently, there is no need to develop a thorough backup plan since all

### 2023-09-24 project-plan.Rmd
removal of dubble

--- 2023-09-21 project-plan.Rmd
@@
 title: "Performance of prediction models for opportunities for improvment in trauma care among women compared to men and across patient cohorts"
-subtitle: ""
 author: "Anna Thoss"
@@
 
-```{=html}
 <!-- The project plan should be 3-4 pages long and written in
@@
 use this document as a template for your own project plan. -->
-```
-```{=html}
+
 <!-- This is an R Markdown document. Markdown is a simple formatting
@@
 details on using R Markdown see http://rmarkdown.rstudio.com -->
-```
-```{=html}
+
 <!-- This is a comment and will not be present in the compiled
@@
 the console and you should only see headings -->
-```
+
 # Introduction
@@
 # References
-
-<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->
+=======

### 2023-09-24 project-plan.Rmd

--- 2023-09-23 project-plan.Rmd
@@
 =======
-
-Introduction
-============
-
-*This is part A) in the project plan [sv: projektredogörelse]. *
-
-*The background/introduction should summarise the scope of the problem,
-what is known about the problem, what is not known, what specific
-knowledge gap the study is supposed to fill and why this is
-important. It should end be stating the aim of the study. The
-background should be 4-5 paragraphs long and each paragraph should be
-between 3-5 sentences. Strive to make the paragraphs about the same
-length.*
-
-*bibliography.bib includes an example reference. Add additional
-references in that file as bibtex entries and then cite in text using:
-"[@exampleKey9999]". One can convert DOIs' to bibtex entries using 
-https://www.doi2bib.org/. If you use Zotero you can simply export 
-entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
-Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*
-
-Methods
-=======
-
-*Refer to the appropriate reporting guideline for details. If you are
-developing, updating or validating a clinical prediction model then
-use
-[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
-you are conducting an observational study, for example a cohort or
-case control study in which you assess associations between some
-exposure and an outcome then use
-[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*
-
-Project Update and Timeline
-===========================
-
-*This is part B) in the project plan [sv: projektredogörelse]. *
-
-*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*
-
-![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)
-
-Back-up Plan
-============
-
-*This is part C) in the project plan [sv: projektredogörelse]*
-
-References
-==========
-
-<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2023-09-25 project-plan.Rmd
inlämningsversion projektredogörelse

--- 2023-09-24 project-plan.Rmd
@@
 editor_options: 
+output:
+  word_document: default
+
   markdown: 
@@
 the console and you should only see headings -->
+```
+## Introduction
 
-# Introduction
+Trauma, defined as the clinical entity composed of the combination of physical injury and the body's associated responses [@gerdin_risk_2015] is a significant public health concern worldwide, accounting for a substantial portion of morbidity and mortality rates, being the cause of 4.3 million fatalities each year [@injuries2020].
+Trauma, and care of patients with traumatic injuries, has long been a significant target point for healthcare improvement programs [@WHO2009].
+Multidisciplinary morbidity and mortality conferences form the cornerstone of initiatives dedicated to enhancing the quality of trauma care and consequently, improving patient outcomes [@Santana2014].
 
-Traumatic injuries are a significant public health concern worldwide,
-accounting for a substantial portion of morbidity and mortality rates,
-being the cause of more than 5 million fatalities each year [@WHO2014].
-Trauma, and care of patients with traumatic injuries, have long been a
-significant target point for healthcare improvement programs [@WHO2009].
-Multidisciplinary mortality and morbidity conferences form the
-cornerstone of initiatives dedicated to enhancing the quality of trauma
-care and, consequently, improving patient outcomes [@Santana2014].
+Audit filters, a system used to flag abnormal clinical parameters as possible opportunities for improvement (OFI), are applied to determine possible events in patient care that are associated with unfavourable outcomes.
+The occurrence of such an event triggers a review process, which in turn leads to a multidisciplinary conference and when appropriate to implementation of corrective measures to address systematic errors and feedback to individual practitioners [@McDermott_1994].
+The use of audit filters has its drawbacks, being associated with high rates of false positives [@Sanddal2011; @Ghorbani2018; @Evans2009].
 
-At mortality and morbidity conferences, representatives from different
-professions and specialties adjacent to trauma care come together to
-discuss care provided to a specific case and compare it to care given
-under optimal conditions. The findings of such a conference are whether
-opportunities for improvement (OFI) can be found in any given case.
-Examples of opportunities for improvement may include lack of resources
-and management errors [@Lazzara2020].
+At mortality and morbidity conferences, representatives from different professions and specialties adjacent to trauma care come together to discuss care provided to a specific patient and compare it to care given under optimal conditions.
+The findings of such a conference are whether OFI can be found in any given case.
+Examples of OFI may include lack of resources and management errors [@Lazzara2020].
 
-Audit filters are applied to determine possible events in patient care
-that are associated with unfavourable outcomes. The occurrence of such
-an event triggers a review process, and when appropriate leads to the
-implementation of corrective measures to address systematic errors and
-feedback to individual practitioners [@McDermott_1994]. The use of audit
-filters has its drawbacks, being associated with high rates of false
-positives. [Saknar tydlig källa här]
+To enhance the precision of the selection process, there have been efforts to introduce trauma mortality prediction models.
+Nevertheless, the performance of these models has been unsatisfactory, being poor predictors for OFI [@Ghorbani2018; @Heim_2016].
+This is likely because these models were originally developed with a focus on predicting mortality rather than morbidity or instances of care failures within the realm of trauma care.
+Machine learning-based prediction models have been found to outperform conventional methods to predict opportunities for improvement [@Attergrim2023].
 
-To enhance the precision of the selection process, there have been
-efforts to introduce trauma mortality prediction models. Nevertheless,
-the performance of these models has been unsatisfactory [@Ghorbani2018]
-[@Heim_2016]. This is likely attributed to the fact that these models
-were originally developed with a focus on predicting mortality rather
-than morbidity or instances of care failures within the realm of trauma
-care. Machine learning-based prediction models have been found to
-outperform conventional methods to predict opportunities for
-improvement. *Hur gör jag med källhänvisning här?, referera till
-preprint artikel?*
+Men make up a disproportionally large group compared to women when it comes to trauma.
+Each year, almost twice as many men die due to traumatic injury when compared to women, with the distribution of different types of injury varying by demographic and region [@WHO2014].
+Females have been shown to have a higher survival following trauma in higher income countries, whereas those in lower income countries show no difference compared to men.
+Differences in cell-mediated immune response following haemorrhage, differences in post-traumatic cytokine levels, and potential protective effects of oestrogen have been proposed to be the cause of the disparity in these groups [@Pendleton2022].
+It is important, that due to the differences between men and women, that the model for prediction has similar performance in these subgroups.
+Furthermore, it is important to establish if trained models perform equally well across different cohorts within the wider span of trauma, such as patients with traumatic brain injury (TBI) and without.
 
-Men make up a disproportionally large group compared to women when it
-comes to trauma. Each year, almost twice as many men die due to
-traumatic injury when compared to women, with the distribution of
-different types of injury varying by demographic and region [@WHO2014].
-It is important, that due to the differences between these groups that
-the model for prediction has similar performance in these subgroups.
-Furthermore, it is important to establish if trained models perform
-equally well across different cohorts within traumatic injuries.
+## Aim
 
-# Aim
+To assess the difference in performance of prediction models for opportunities for improvement in trauma care among women compared to men and across clinical trauma cohorts.
 
-To assess the difference in performance of prediction models for
-opportunities for improvement in trauma care among women compared to men
-and across clinical trauma cohorts.
+## Methods
 
-# Methods
+#### Study setting and participants
 
-This is a registry-based study, encompassing all trauma patients
-enlisted in both the Karolinska University Hospital trauma registry and
-the trauma care quality database spanning from 2012 to 2022. The
-Karolinska University Hospital in Solna is equivalent to a level 1
-trauma centre and manages approximately 1,500 acute trauma patients each
-year[@Ghorbani2018][Swetrau, rapport]. The trauma registry comprises all
-patients admitted to the hospital with trauma team activation,
-irrespective of their Injury Severity Score (ISS), as well as patients
-with an ISS exceeding 9. Included in the registry is data on vital
-signs, injuries, interventions, times, and patient demographics.
+We will conduct a registry-based study, encompassing all trauma patients enlisted in both the Karolinska University Hospital trauma registry and the trauma care quality database spanning from 2012 to 2022.
+The Karolinska University Hospital in Solna is equivalent to a level 1 trauma centre and manages approximately 1,500 acute trauma patients each year [@Ghorbani2018; @swetrau].
+The trauma registry comprises all patients admitted to the hospital with trauma team activation, irrespective of their Injury Severity Score (ISS), as well as patients with an ISS exceeding 9.
+Included in the registry is data on vital signs, injuries, interventions, times, and patient demographics.
 
-The trauma registry encompasses approximately 12,000 patients treated
-from 2012 to 2022. The trauma care quality database in turn is a subset
-of the trauma registry, including about 6000 patients selected for
-review in the years 2013-2022.
+The trauma registry encompasses approximately 12,000 patients treated from 2012 to 2022.
+The trauma care quality database in turn is a subset of the trauma registry, including about 6000 patients selected for review in the years 2013-2022.
 
-The outcome of this study is opportunities for improvement (OFI), which
-are identified through meetings of the multidisciplinary review board
-during mortality and morbidity conferences. An OFI is a consensus
-decision made during a conference and includes proposals for corrective
-actions. These in turn are recorded in the trauma care quality database.
-The trauma care quality database also includes data relevant to
-mortality and morbidity conferences, such as audit filters. The
-mortality and morbidity conferences at Karolinska University Hospital
-involve professionals from various specialties within trauma care,
-including surgery, neurosurgery, orthopedics, anesthesia, intensive
-care, nursing, and radiology. Cases considered for a conference go
-through multiple rounds of review. Mortality leads to direct inclusion.
-Between 2013 and 2017, non-fatal poor outcomes were identified through
-reviews conducted by a specialized trauma nurse to identify potential
-OFI. Starting in 2017, the process was formalized, involving an initial
-review by a specialized trauma nurse along with the application of audit
-filters. All cases that were flagged either during the initial nurse
-review or by the audit filters underwent a secondary review conducted by
-two specialized nurses. If the second review identified potential areas
-for improvement (OFI), the case was marked for discussion in a
-multidisciplinary conference.
+OFI are identified through meetings of the multidisciplinary review board during mortality and morbidity conferences.
+An OFI is a consensus decision made during a conference and includes proposals for corrective actions.
+These in turn are recorded in the trauma care quality database.
+The trauma care quality database also includes data relevant to mortality and morbidity conferences, such as audit filters.
+The mortality and morbidity conferences at Karolinska University Hospital involve professionals from various specialties within trauma care, including surgery, neurosurgery, orthopedics, anesthesia, intensive care, nursing, and radiology.
+Cases considered for a conference go through multiple rounds of review.
+Mortality leads to direct inclusion.
+Between 2013 and 2017, non-fatal poor outcomes were identified through reviews conducted by a specialized trauma nurse to identify potential OFI.
+Starting in 2017, the process was formalized, involving an initial review by a specialized trauma nurse along with the application of audit filters.
+All cases that were flagged either during the initial nurse review or by the audit filters underwent a secondary review conducted by two specialized nurses.
+If the second review identified potential areas for improvement (OFI), the case was marked for discussion in a multidisciplinary conference.
 
-All patients reviewed for possible opportunities for improvement between
-the years 2013 and 2021 are included in the study, with the exception of
-patients under the age of 15 due to differing clinical pathways in this
-subgroup.
+![](assets/OFI-flow.png)
+Figure illustrating process of mortality and morbidity conference at Karolinska University Hospital.
 
-The outcome of this study will be defined as "Yes, there is a
-significant difference in models prediction of OFI in different cohorts"
-or "No, there is no significant difference in models prediction of OFI
-in different cohorts." Data for this will be extracted from the trauma
-care quality database and run through different previously trained
-machine learning models. *osäker på om outcome här är hållbart
-formulerad*
+#### Eligibility criteria
 
-All variables from the trauma registry will be considered as potential
-predictors in our analysis. These variables include information spanning
-the pre-hospital, initial care, and subsequent in-hospital phases, which
-include initial vital signs, timing, and types of procedures and
-interventions, duration and level of care, injury details, mechanisms of
-injury, injury types, and standard demographic information. This array
-of predictors comprises both continuous and categorical variables,
-ultimately resulting in the final models utilizing a total of 45
-predictors. (infoga bilaga) All available data will be used, with a
-sample size of 6310 patients. This is due to previously developed
-prediction models being based on all available data, as well as
-including all 45 predictors regardless of the learner used for the
-development of the model.
+All patients reviewed for possible OFI between the years 2013 and 2022 are included in the study, with the exception of patients under the age of 15 due to differing clinical pathways in this subgroup.
 
-The models have been developed and trained on the data set previously to
-this study. Eight models are included, all being developed using the
-Tidymodels framework (). These include logistic regression (LR), random
-forest (RF), decision tree (DT), support vector machine with a radial
-basis kernel (SVM), XGBoost, LightGBM, CatBoost, and k-nearest neighbor
-(k-NN).
+#### Outcome
 
-To assess different models' performance across different cohorts, data
-from different cohorts run through multiple models. These cohorts are
-defined as follows:
+The outcome of this study will be defined as "Yes, there is a significant difference in models prediction of OFI in different cohorts" or "No, there is no significant difference in models prediction of OFI in different cohorts." Data for this will be extracted from the trauma care quality database and run through different previously trained machine learning models.
 
-1.  Blunt multi-system trauma: Blunt trauma with AIS ≥ 3 in at least two
-    of the following AIS body regions: head, face, neck, thorax,
-    abdomen, spine, or upper and lower extremities.
+#### Predictors
 
-2.  Isolated severe TBI: Injury isolated to the area of the brain with
-    AIS \> 2 and:
+All variables from the trauma registry will be considered as potential predictors in our analysis and have already been incorporated into the models.<!-- Note that the models are already developed.
+So you should write about what have already been done,i.e.
+these variables are already incorporated into the
+models-->These variables include information spanning the pre-hospital, initial care, and subsequent in-hospital phases, which include initial vital signs, timing, and types of procedures and interventions, duration and level of care, injury details, mechanisms of injury, injury types, and standard demographic information.
+This array of predictors comprises both continuous and categorical variables, ultimately resulting in the final models utilizing a total of 45 predictors.
+
+All available data will be used, with a sample size of 6310 patients.
+This is due to previously developed prediction models being based on all available data, as well as including all 45 predictors regardless of the learner used for the development of the model.
+
+The models have been developed and trained on the data set previous to this study [@Attergrim2023].
+Eight models are included, all being developed using the Tidymodels framework [@tidymodels].
+These include logistic regression (LR), random forest (RF), decision tree (DT), support vector machine with a radial basis kernel (SVM), XGBoost, LightGBM, CatBoost, and k-nearest neighbor (k-NN).
+
+#### Cohorts
+
+The different models' performance were assessed in the following cohorts:
+
+1.  Isolated severe TBI: Injury isolated to the area of the brain with AIS \> 2 and:
 
@@
 
-3.  Blunt multi-system trauma without TBI: Blunt trauma with AIS \> 2
-    and injuries in at least two of the following AIS body regions:
-    head, face, neck, thorax, abdomen, spine, or upper and lower
-    extremities and no injury to the brain with AIS \>2 and:
+2.
+Blunt multi-system trauma without TBI: Blunt trauma with AIS \> 2 and injuries in at least two of the following AIS body regions: head, face, neck, thorax, abdomen, spine, or upper and lower extremities and no injury to the brain with AIS \>2 and:
 
@@
 
-4.  Blunt multi-system trauma with TBI: Blunt trauma with AIS \> 2 and
-    injuries in at least two of the following AIS body regions: head,
-    face, neck, thorax, abdomen, spine, or upper and lower extremities
-    and with injury to the brain with AIS \>2 and:
+3.
+Blunt multi-system trauma with TBI: Blunt trauma with AIS \> 2 and injuries in at least two of the following AIS body regions: head, face, neck, thorax, abdomen, spine, or upper and lower extremities and with injury to the brain with AIS \>2 and:
 
@@
 
-Along with these cohorts, all machine learning models will be assessed
-in regard to differences between male and female gender in their
-prediction of OFI.
+Along with these cohorts, all machine learning models will be assessed in regard to differences between male and female sex in their prediction of OFI.
 
-We will employ a 95% confidence level and a 5% significance level for
-our analysis, with all statistical analysis conducted using R.
+#### Statistical analysis
 
-# Project Update and Timeline
+We will employ a 95% confidence level and a 5% significance level for our analysis, with all statistical analysis conducted using R.
 
-![**The Gantt chart will be displayed here when you place it in the
-assets folder under "project-gantt.png"**](assets/project-gantt.png)
+## Ethical considerations
 
-Gantt chart over the project timeline.
+The nature of this study is non-invasive.
+However, handling of registry data will constitute a risk for breach of patients integrity.
+The registry used in this study does contain personal identification.
+In order to minimize the risk of breach, patients have been anonymized after collection of data and been assigned a new identification number upon entry into dataset.
+Due to the nature of the patient population, patients have not consented previous to entry into the registry, thus, there is an opt out principle applied here.
+This is common with registry data in Sweden, and opens up opportunities for studies.
+In turn, more scientific knowledge has great potential for benefit, both on the societal scale, as well as the individual.
+Thus, the risk for the individual is compatibly small.
+Further, no data is presented of individual patients in the study, mitigating risk of violating personal integrity.
+The study has been granted ethical permit, with the record number 2021-02541 and 2021-03531.
 
-# Back-up Plan
+## Project Update and Timeline
 
-Currently, there is no need to develop a thorough backup plan since all
-the necessary data for the statistical analysis has been collected and
-is accessible.
+As of the date of submission, all data has been collected, and is accessible in the database.
+All machine learners have been trained on the data.
+I have spent the last weeks on reading up as well as learning the fundamentals of the coding that I will do later on.
+
+Currently I am in the process of going through the data, and will be sorting it by cohort and run the models on different cohorts.
+This will take a significant portion of the next weeks.
+For this, some code will be written.
+Following this, the statistical analysis will be performed.
+Here different models will be compared and assessed in their effectiveness in predicting OFI.
+
+Currently, we see no need to develop a thorough backup plan since all the necessary data for the statistical analysis has been collected and is accessible.
+However, some cohorts might be redefined or added, changing the direction of the project.
+
+![](assets/project-gantt.png)
+
+Gantt chart overview of the project timeline.
 
 # References
-=======

### 2023-12-04 manuscript.Rmd
mm + res

@@
 ---
-title: "Manuscript title here"
-author: "Your name here"
+title: "Performance of prediction models for opportunities for improvment in trauma care among women compared to men and across patient cohorts"
+author: "Anna Thoss"
 output:
@@
 
+```{=html}
 <!-- This is an R Markdown document. Markdown is a simple formatting
@@
 the console and you should only see headings -->
-
+```
+```{=html}
 <!-- Structure and most comments are from the Strengthening the
@@
 Epidemiology (STROBE):explanation and elaboration.  -->
-
+```
+```{=html}
 <!-- Indicate the study’s design with a commonly used term in the
@@
 study" -->
+```
+# Abstract
 
-Abstract
-========
-
+```{=html}
 <!-- Provide in the abstract an informative and balanced summary of
 what was done and what was found. Not more than 300 words. -->
+```
+## Background
 
-Background
-----------
+## Methods
 
-Methods
--------
+## Results
 
-Results
--------
+## Conclusion
 
-Conclusion
-----------
+# Introduction
 
-Introduction
-============
+Trauma, defined as the clinical entity composed of the combination of physical injury and the body's associated responses [@gerdin_risk_2015] is a significant public health concern worldwide, accounting for a substantial portion of morbidity and mortality rates, being the cause of 4.3 million fatalities each year [@injuries2020].
+It is the leading cause of death for people aged 18-39 [@Peng2023-gd].
+Trauma, and care of patients with traumatic injuries, have long been a significant target point for healthcare improvement programs [@WHO2009].
+Multidisciplinary morbidity and mortality conferences form a cornerstone of initiatives dedicated to enhancing the quality of trauma care and consequently, improving patient outcomes [@Santana2014].
 
-<!-- Explain the scientific background and rationale for the
-investigation being reported. End by stating the aim or specific
-objectives, including any prespecified hypotheses. The introduction
-should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
-6-8 lines long. Support every statement with a citation -->
+#### Opportunities for improvement
 
-You can cite document here like this [@exampleKey9999]. Open the file
-bibliography.bib to learn more.
+Audit filters, a system used to flag abnormal clinical parameters as possible opportunities for improvement (OFI), are applied to determine possible events in patient care that are associated with unfavourable outcomes.
+The occurrence of such an event triggers a review process, which in turn leads to a meeting of the multidisciplinary review board during mortality and morbidity conferences.
+An OFI is a consensus decision made during a conference and includes proposals for corrective actions, and when appropriate implementation of corrective measures to address systematic errors and feedback to individual practitioners [@McDermott_1994].
+At mortality and morbidity conferences, representatives from different professions and specialities adjacent to trauma care come together to discuss care provided to a specific patient and compare it to care given under optimal conditions.
+The findings of such a conference are whether OFI can be found in any given case.
+Examples of OFI may include lack of resources and management errors [@Lazzara2020].
+The use of audit filters has its drawbacks, being associated with high rates of false positives [@Sanddal2011; @Ghorbani2018; @Evans2009].
 
-Methods 
-=======
+#### Artificial intelligence
 
-Study design
-------------
+In recent years, artificial intelligence (AI) has made great advances, making it a topic of interest in regard to healthcare applications.
+Research on AI implementation for clinical use has become commonplace and has already been tried in some clinical contexts, such as aiding in the diagnosis of pathologies through analysis of radiological images [@Cellina2022-qp] as well as epidemiological prediction of disease patterns, an example being during the COVID-19 pandemic [@Chen2020-ue].
+
+With AI, computing is used to mimic human intelligence, characterized by its ability to learn and reason.
+Through predictive modelling, AI can be used as a tool to problem solve.
+Computation has the advantage of analysing great quantities of data in a short time, giving it an edge [@Cellina2022-qp].
+The primary way to achieve AI is through machine learning (ML), in which statistical modelling is applied to a computer system where it learns from previous data, and thus circumvents human programming.
+The approach with ML has increasingly been utilized to analyse data and can be used to gain further information from data, such as making predictions of outcomes[@MLJung].
+The ability to improve modelling algorithms with the use of ML thus has the ability to improve said algorithms autonomously, which in turn opens the door for a wide array of applications in healthcare.
+One such application is to develop more accurate prediction models that can further quality improvement programs.
+
+### AI in trauma care
+
+To enhance the precision of the selection process for mortality and morbidity conferences, there have been efforts to introduce trauma mortality prediction models.
+Nevertheless, the performance of these models has been unsatisfactory, being poor predictors for OFI [@Ghorbani2018; @Heim_2016].
+This is likely because these models were originally developed with a focus on predicting mortality rather than morbidity or instances of care failures within the realm of trauma care.
+
+Machine learning-based prediction models have been found to outperform conventional methods to predict opportunities for improvement [@Attergrim2023].
+
+#### Trauma care in different patient cohorts
+
+Men make up a disproportionally large group compared to women when it comes to trauma.
+Each year, almost twice as many men die due to traumatic injury when compared to women, with the distribution of different types of injury varying by demographic and region [@WHO2014].
+Females have been shown to have a higher survival following trauma in higher income countries, whereas those in lower income countries show no difference compared to men.
+Differences in cell-mediated immune response following haemorrhage, differences in post-traumatic cytokine levels, and potential protective effects of oestrogen have been proposed to be the cause of the disparity between these groups [@Pendleton2022].
+It is important, that due to the differences between men and women, that the model for prediction has similar performance in these subgroups.
+Furthermore, it is important to establish if trained models perform equally well across different cohorts within the wider span of trauma, such as patients with traumatic brain injury (TBI) and without.
+
+## Aim
+
+To assess the difference in performance of prediction models for opportunities for improvement in trauma care among women compared to men and across clinical trauma cohorts.
+
+# Methods
+
+## Methods
+
+#### Study setting and participants
+
+We conducted a registry-based study, encompassing all trauma patients enlisted in both the Karolinska University Hospital trauma registry and the trauma care quality database spanning from 2012 to 2022.
+The Karolinska University Hospital in Solna is equivalent to a level 1 trauma centre and manages approximately 1,500 acute trauma patients each year [@Ghorbani2018; @swetrau].
+The trauma registry comprises all patients admitted to the hospital with trauma team activation, irrespective of their Injury Severity Score (ISS), as well as patients with an ISS exceeding 9.
+Included in the registry is data on vital signs, injuries, interventions, times, and patient demographics.
+
+The trauma registry encompasses approximately 12,000 patients treated from 2012 to 2022.
+The trauma care quality database in turn is a subset of the trauma registry, including about 6000 patients selected for review in the years 2013-2022.
+
+OFI are identified through meetings of the multidisciplinary review board during mortality and morbidity conferences.
+An OFI is a consensus decision made during a conference and includes proposals for corrective actions.
+These in turn are recorded in the trauma care quality database.
+The trauma care quality database also includes data relevant to mortality and morbidity conferences, such as audit filters.
+The mortality and morbidity conferences at Karolinska University Hospital involve professionals from various specialities within trauma care, including surgery, neurosurgery, orthopaedics, anaesthesia, intensive care, nursing, and radiology.
+Cases considered for a conference go through multiple rounds of review.
+Mortality leads to direct inclusion.
+
+Between 2013 and 2017, non-fatal poor outcomes were identified through reviews conducted by a specialized trauma nurse to identify potential OFI.
+Starting in 2017, the process was formalized, involving an initial review by a specialized trauma nurse along with the application of audit filters.
+All cases that were flagged either during the initial nurse review or by the audit filters underwent a secondary review conducted by two specialized nurses.
+If the second review identified potential areas for improvement (OFI), the case was marked for discussion in a multidisciplinary conference.
+
+![](assets/OFI-flow.png) Figure illustrating process of mortality and morbidity conference at Karolinska University Hospital.
+
+#### Eligibility criteria
+
+All patients reviewed for possible OFI between the years 2013 and 2022 are included in the study, with the exception of patients under the age of 15 due to differing clinical pathways in this subgroup.
+
+#### Outcome
+
+The outcome of this study will be defined as "Yes, there is a significant difference in models prediction of OFI in different cohorts" or "No, there is no significant difference in models prediction of OFI in different cohorts." Data for this will be extracted from the trauma care quality database and run through a number of machine learning models, that have been trained previous to this study.
+
+#### Predictors
+
+All variables from the trauma registry are considered potential predictors in our analysis and have been incorporated into the models.
+<!-- Note that the models are already developed.
+So you should write about what have already been done,i.e.
+these variables are already incorporated into the
+models-->These variables include information spanning the pre-hospital, initial care, and subsequent in-hospital phases, which include initial vital signs, timing, and types of procedures and interventions, duration and level of care, injury details, mechanisms of injury, injury types, and standard demographic information.
+This array of predictors comprises both continuous and categorical variables, ultimately resulting in the final models utilizing a total of 45 predictors.
+
+All available data was used, with a sample size of 6310 patients.
+This is due to previously developed prediction models being based on all available data, as well as including all 45 predictors regardless of the learner used for the development of the model.
+
+The models have been developed and trained on the data set previous to this study [@Attergrim2023].
+Eight models are included, all being developed using the Tidymodels framework [@tidymodels].
+These include logistic regression (LR), random forest (RF), decision tree (DT), support vector machine with a radial basis kernel (SVM), XGBoost, LightGBM, CatBoost, and k-nearest neighbor (k-NN).
+
+#### Cohorts
+
+The different models' performance were assessed in the following cohorts:
+
+1.  Isolated severe TBI: Injury isolated to the area of the brain with AIS \> 2 and:
+
+    a.  pre-hospital GCS of \< 9
+
+        or
+
+    b.  Pre- or in-hospital intubation.
+
+2.
+Blunt multi-system trauma without TBI: Blunt trauma with AIS \> 2 and injuries in at least two of the following AIS body regions: head, face, neck, thorax, abdomen, spine, or upper and lower extremities and no injury to the brain with AIS \>2 and:
+
+    a.  pre-hospital GCS of \< 9
+
+        or
+
+    b.  Pre- or in-hospital intubation.
+
+3.
+Blunt multi-system trauma with TBI: Blunt trauma with AIS \> 2 and injuries in at least two of the following AIS body regions: head, face, neck, thorax, abdomen, spine, or upper and lower extremities and with injury to the brain with AIS \>2 and:
+
+    a.  pre-hospital GCS of \< 9
+
+        or
+
+    b.  Pre- or in-hospital intubation
+
+Along with these cohorts, all machine learning models were assessed regarding differences between male and female sex in their prediction of OFI.
+
+#### Statistical analysis
+
+R was employed for all statistical analysis.
+For the analysis, a 95% confidence level and 5% significance level were employed.
+For each model, the performance was assessed in the predefined cohorts and compared to other models performance in regards to sensitivity and specificity.
+
+## Study design
 
@@
 
-Setting
--------
+## Setting
 
+```{=html}
 <!-- Describe the setting, locations, and relevant dates, including
 periods of recruitment, exposure, follow-up, and data collection -->
+```
+## Participants
 
-Participants
-------------
-
+```{=html}
 <!-- Cohort study: Give the eligibility criteria, and the sources and
@@
 exposed and unexposed -->
-
+```
+```{=html}
 <!-- Case-control study: Give the eligibility criteria, and the
@@
 of controls per case -->
-
+```
+```{=html}
 <!-- Cross-sectional study: Give the eligibility criteria, and the
 sources and methods of selection of participants -->
+```
+## Variables and data sources/measurements
 
-Variables and data sources/measurements
----------------------------------------
-
+```{=html}
 <!-- Clearly define all outcomes, exposures, predictors, potential
@@
 of assessment methods if there is more than one group -->
-
-Bias 
-----
+```
+## Bias
 
@@
 
-Study size 
-----------
+## Study size
 
@@
 
-Quantitative variables
-----------------------
+## Quantitative variables
 
+```{=html}
 <!-- Explain how quantitative variables were handled in the
@@
 -->
+```
+## Statistical methods
 
-Statistical methods
--------------------
-
+```{=html}
 <!--
@@
 -->
+```
+# Results
 
-Results
-=======
+### Participants
 
+Karolinska trauma registry contains 11864 patients, of which 6310 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
+Through this, 1316 patients were selected for the morbidity conference, and 596 patients for the mortality conference.
+Out of 596 deaths reviewed, 35 were found to be preventable (n=4) or possibly preventable (n=31).
+The remaining 561 patients were thus rendered non-preventable without OFI.
+Further, 1316 alive patients were included during a morbidity conference, of which 396 (7%) were identified as cases with OFI.
+
+Among the 6310 patients, the majority (n=4383) (69%) were male, and the mean age was 45 (SD 21). Overall, the mortality in the group was 599 (9%).
+
+In the group identified as OFI (n = 431), the mean age was somewhat higher than in the group without OFI (n=5879) (mean 48 vs 45 years).
+For OFI patients, they were most commonly treated at an intensive care unit (35%), whereas patients without OFI were most commonly treated at a general ward (37%).
+ISS was also higher for OFI patients (mean: 19, SD:11 vs mean: 12, SD:13).
+In-hospital intubation occurred in 74 (17%) OFI patients, compared to non-OFI (n=460, 8%).
+They had a longer time till definitive treatment (median: 143 vs 99 minutes from hospital arrival).
+Definitive treatments differed, with OFI patients receiving more interventions, particularly radiological (7% vs 1%).
+
+The variables most frequently missing data were "Time to normal BE" (n=6107), "Pre-hospital intubation type" (N=5812), and "Emergency department intubation type" (n=5749).
+
+### Descriptive data
+
+### Outcome data
+
+### Main result
+
+### Other analysis
+
+```{=html}
 <!--
@@
 -->
-
+```
 You can include code in this document like this:
@@
 
-Discussion
-==========
+# Discussion
 
+```{=html}
 <!--
@@
 -->
-
-Conclusion
-==========
+```
+# Conclusion
 
@@
 
-References
-==========
+# References
 

### 2023-12-04 project-plan.Rmd
mm + res

@@
 Females have been shown to have a higher survival following trauma in higher income countries, whereas those in lower income countries show no difference compared to men.
-Differences in cell-mediated immune response following haemorrhage, differences in post-traumatic cytokine levels, and potential protective effects of oestrogen have been proposed to be the cause of the disparity in these groups [@Pendleton2022].
+Differences in cell-mediated immune response following haemorrhage, differences in post-traumatic cytokine levels, and potential protective effects of oestrogen have been proposed to be the cause of the disparity between these groups [@Pendleton2022].
 It is important, that due to the differences between men and women, that the model for prediction has similar performance in these subgroups.
@@
 
-Along with these cohorts, all machine learning models will be assessed in regard to differences between male and female sex in their prediction of OFI.
+Along with these cohorts, all machine learning models will be assessed regarding differences between male and female sex in their prediction of OFI.
 

### 2023-12-04 functions/select_varaibles.R

@@ -1,0 +1,10 @@
+#' Select Varaibles
+#'
+#' Write a short description of your function here 
+#' @param x Describe argument x here 
+#' @export 
+select_varaibles <- function(x) {
+    ## Replace with the contents of your function
+    y <- x + 1
+    return (y)
+}

### 2023-12-08 manuscript.Rmd
Comment on manuscript

@@
 
+<!-- Expand on this section to introduce AI/ML in trauma care more generally, and then focus on the application of ML to select patients for M&M -->
+
 To enhance the precision of the selection process for mortality and morbidity conferences, there have been efforts to introduce trauma mortality prediction models.
@@
 #### Trauma care in different patient cohorts
+
+<!-- Similar performance across different subgroups is also important from a fairness perspective, meaning that if the model performs worse among women, then we risk providing lower quality care to women, because we might miss important OFI -->
 
@@
 The trauma care quality database in turn is a subset of the trauma registry, including about 6000 patients selected for review in the years 2013-2022.
+
+<!-- Consider citing Hussein's paper in your definition of OFI: "The outcome was the presence of OFI, as decided by the morbidity and mortality conference.
+An OFI is any failure of care including, but not limited to, any potentially preventable or preventable death, delay in treatment, clinical judgment error, missed diagnosis and technical error as decided by the mortality and morbidity conference.
+The study outcome is binary with the levels “Yes - At least one OFI identified” and “No - No OFI identified”."
+
+https://sjtrem.biomedcentral.com/articles/10.1186/s13049-023-01157-y
+
+-->
 
@@
 
-Karolinska trauma registry contains 11864 patients, of which 6310 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
+The Karolinska trauma registry contains 11864 patients, of which 6310 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
 Through this, 1316 patients were selected for the morbidity conference, and 596 patients for the mortality conference.

### 2023-12-11 manuscript.Rmd

--- 2023-12-04 manuscript.Rmd
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r setup, include=FALSE, echo=FALSE}
+library(knitr)
+library(kableExtra)
+library(flextable)
+
+knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = NA)
+
+set_flextable_defaults(
+  font.size = 8,
+  theme_fun = theme_vanilla,
+  padding.top = 1,
+  padding.bottom = 1,
+  padding.left = 1,
+  padding.right = 1
+)
+
+output.format <- knitr::opts_knit$get('rmarkdown.pandoc.to')
+
+source("./main.R")
 ```
@@
 
+```{r tab1, echo=FALSE, fig.cap="Selected sample characteristics", fig.align='center' }
+if (output.format == "latex") {
+  tab1 <- t1kable(create.tableone(dataset), "latex", longtable = T, booktabs = T) %>%
+    kable_styling(latex_options = c("repeat_header", "condensed"), repeat_header_text = "(cont.)", font_size = 8) %>%
+    kableExtra::footnote(general = "\\\\textit{Definition of  abbreviations:} OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\\\\newline Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.", general_title = "",threeparttable =  TRUE, escape = FALSE)
+} else if (output.format == "docx") {
+    tab1 <- t1flex(create.tableone(dataset)) %>%
+      add_footer_lines("OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\n Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.") %>%
+      autofit()
+}
+
+tab1
+```
+
 Karolinska trauma registry contains 11864 patients, of which 6310 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
@@
 ### Main result
+
+```{r main.result, echo=FALSE}
+kable(cohort.performance.results, caption = "LÄGG TILL CAPTION HÄR")
+```
+
+<!-- KS: P-value was < 10-6 between the cohorts. Present that value where it suits you -->
+
+```{r roc.curves, out.width="80%",fig.cap="ROC Curves. LÄGG CAPTION HÄR", echo=FALSE, fig.align='center'}
+knitr::include_graphics("figures/roc.pdf", error = FALSE)
+```
 

### 2023-12-11 project-plan-saved.Rmd

--- 2023-09-24 project-plan.Rmd
@@
 the console and you should only see headings -->
-
+```
 # Introduction
 
-Traumatic injuries<!-- Call it just trauma--> <!-- KS: when removing the word injuries it can be good to define "trauma" to differ from psychological trauma,
-A good citation is:
-@book{gerdin_risk_2015,
-    location = {Stockholm},
-    title = {The risk of dying: predicting trauma mortality in urban Indian
-             hospitals},
-    isbn = {978-91-7676-157-1},
-    shorttitle = {The risk of dying},
-    publisher = {Karolinska Institutet},
-    author = {Gerdin, Martin},
-    date = {2015},
-    note = {{OCLC}: 942942790},
-}
-A good read if you have time ;), PDF URL: https://openarchive.ki.se/xmlui/bitstream/handle/10616/44832/Thesis_Martin_Gerdin.pdf?sequence=3&isAllowed=y
---> 
-are a significant public
-health concern worldwide, accounting for a substantial portion of
-morbidity and mortality rates, being the cause of more than 5 million
-fatalities each year [@WHO2014]<!-- This is an old reference, most
-current figure is 4.3 million. Use:
-@article{injuries2020,
-  title =	 {Injuries—Level 1 cause},
-  volume =	 {396},
-  ISSN =	 {0140-6736},
-  journal =	 {The Lancet},
-  publisher =	 {Elsevier BV},
-  author =	 {{GBD 2019 Diseases and Injuries Collaborators}},
-  year =	 {2020},
-  month =	 {Oct},
-}
--->.  Trauma, and care of patients
-with traumatic injuries, have long been a significant target point for
-healthcare improvement programs [@WHO2009].  Multidisciplinary
-mortality and morbidity <!-- KS: The order "morbidity and mortality" is mainly used in the literature. I also received a comment on this in my final version. --> conferences form the cornerstone of
-initiatives dedicated to enhancing the quality of trauma care and,
-consequently, improving patient outcomes [@Santana2014].
+Traumatic injuries are a significant public health concern worldwide,
+accounting for a substantial portion of morbidity and mortality rates,
+being the cause of more than 5 million fatalities each year [@WHO2014].
+Trauma, and care of patients with traumatic injuries, have long been a
+significant target point for healthcare improvement programs [@WHO2009].
+Multidisciplinary mortality and morbidity conferences form the
+cornerstone of initiatives dedicated to enhancing the quality of trauma
+care and, consequently, improving patient outcomes [@Santana2014].
 
@@
 implementation of corrective measures to address systematic errors and
-feedback to individual practitioners [@McDermott_1994]. The use of
-audit filters has its drawbacks, being associated with high rates of
-false positives. [Saknar tydlig källa här]<!-- Consider changing the
-order of these paragraphs, so that you introduce audit filters before
-mortality and morbidity conferences, considering that audit filters
-are used to select patients for those conferences-->
-<!-- KS: you will probably need to explain exactly what audit fitlers are.
-That they are a system that flags abnormal clinical parameters as possible OFI.
-These flag cases are then reviewed by two nurses and then a conference -->
+feedback to individual practitioners [@McDermott_1994]. The use of audit
+filters has its drawbacks, being associated with high rates of false
+positives. [Saknar tydlig källa här]
 
@@
 # Methods
-<!-- I would find this section easier to read if you followed some
-reporting guideline, like TRIPOD--> <!-- KS: prefer "We conducted a registry[...]" -->This is a registry-based study,
-encompassing all trauma patients enlisted in both the Karolinska
-University Hospital trauma registry and the trauma care quality
-database spanning from 2012 to 2022. The Karolinska University
-Hospital in Solna is equivalent to a level 1 trauma centre and manages
-approximately 1,500 acute trauma patients each
-year[@Ghorbani2018][Swetrau, rapport]. The trauma registry comprises
-all patients admitted to the hospital with trauma team activation,
+
+This is a registry-based study, encompassing all trauma patients
+enlisted in both the Karolinska University Hospital trauma registry and
+the trauma care quality database spanning from 2012 to 2022. The
+Karolinska University Hospital in Solna is equivalent to a level 1
+trauma centre and manages approximately 1,500 acute trauma patients each
+year[@Ghorbani2018][Swetrau, rapport]. The trauma registry comprises all
+patients admitted to the hospital with trauma team activation,
 irrespective of their Injury Severity Score (ISS), as well as patients
@@
 
-All patients reviewed for possible opportunities for improvement <!-- KS: use OFI consistently after you have abbreviated it --> between
+All patients reviewed for possible opportunities for improvement between
 the years 2013 and 2021 are included in the study, with the exception of
@@
 
-<!-- KS: Gianluigi will most likely not like this explaination so if you have time just come up with a random backup project -->
 Currently, there is no need to develop a thorough backup plan since all
@@
 # References
-=======
+
+<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2023-12-11 project-plan.Rmd
pre analysis

@@
 Examples of OFI may include lack of resources and management errors [@Lazzara2020].
+
+!Here, add explanation of what machine learning is, along with what has been done in regards to ML and trauma care thus far!
 
@@
 The registry used in this study does contain personal identification.
-In order to minimize the risk of breach, patients have been anonymized after collection of data and been assigned a new identification number upon entry into dataset.
+In order to minimize the risk of breach, patients have been semi-anonymized after collection of data and been assigned a new identification number upon entry into dataset.
 Due to the nature of the patient population, patients have not consented previous to entry into the registry, thus, there is an opt out principle applied here.

### 2023-12-11 main.R

@@ -80,3 +80,17 @@ bm.no.tbi <- dataset[dataset$cohort == "blunt multisystem without TBI",]
 
 bm.no.tbi.auc <- pROC::auc(ifelse(bm.no.tbi$ofi == "Yes", 1, 0), bm.no.tbi$repeat_1)
 print(bm.no.tbi.auc)
+
+#### Table1
+
+install.packages("gtsummary")
+
+library(tidyverse)
+library(gtsummary)
+
+table1 <- 
+  trial %>%
+  tbl_summary(include = c(age, grade, response))
+
+  
+

### 2023-12-11 main.R
attempt 2 solving merge conflict

@@ -18,42 +18,10 @@
 ## them). They all need to be sourced however when you compile your
 ## manuscript file or run this file as a job, as that happens in a
 ## clean R session.
-library(rofi)
-noacsr::source_all_functions()
 
-## Import data
-data <- import_data(test = TRUE) # Need to add test = TRUE when you run
-
-combined.datasets <- rofi::merge_data(data,test = TRUE)
-combined.datasets$ofi <- rofi::create_ofi(combined.datasets)
-
-# I suggest switching to create.dataset()  function when moving to real data, 
-# it adds 2022 data but it doesent work for the test data?
-# It does everything from import_data to clean_all_predictors
-
-
-##############################################
-# JA: Added functions from previous projects #
-##############################################
-
-#If you run create.data() you can skip the steps untill i mark it
-combined.datasets <- clean_audit_filters(combined.datasets)
-
-## Separate and store cases without known outcome
-missing.outcome <- is.na(combined.datasets$ofi)
-combined.datasets <- combined.datasets[!missing.outcome,]
-
-# We exkluder patients under 15 since they go through a different clinical and review pathway.
-combined.datasets <- combined.datasets[combined.datasets$pt_age_yrs > 14,]
-
-## Fix formating and remove wrong values like 999
-combined.datasets <- clean_all_predictors(combined.datasets)
-
-# Skip untill the next step
 set.seed(2023) # for reproducibility
 dir.create("results", showWarnings = FALSE) # ensure results dir exists
 
-dataset <- combined.datasets
 library(rofi)
 library(tidymodels)
 library(gmodels)
@@ -61,22 +29,13 @@ library(dplyr)
 library(pROC)
 library(ggplot2)
 
-# Adds specific trauma quality cohorts
-dataset <- create_cohorts(dataset)
 noacsr::source_all_functions()
 tidymodels_prefer()
 
-# Adds "other cohort" instead of NA
-dataset$cohort[is.na(dataset$cohort) == TRUE] <- "Other cohort"
 # Settings
 n.bootstraps <- 1000
 model.name <- "xgb"
 
-# You can visualize the cohorts through: table(dataset$cohort)
-########################
-# End of edits from JA #
-########################
[…]

### 2023-12-12 manuscript.Rmd
attempt 2 solving merge conflict
Update table 1 and save as word file

@@
   html_document: default
+  word_document: default
 bibliography: bibliography.bib
@@
     tab1 <- t1flex(create.tableone(dataset)) %>%
-      add_footer_lines("OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\n Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.") %>%
+      add_footer_lines("OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.") %>%
       autofit()

### 2023-12-12 functions/table_1.R

@@ -303,14 +303,13 @@ format.table1 <- function(data) {
     dt_alarm_scene ="Injury scene time",        
     dt_ed_norm_be ="Time to normal base excess",
     NumberOfActions ="Number of interventions",
-    NumberOfInjuries ="Number of injuries") ##Requires library(labelled)
+    NumberOfInjuries ="Number of injuries",
+    cohort = "Clinical cohort") ##Requires library(labelled)
   
   return(data)
 }
 
 create.tableone <- function(data) {
-  data <- subset(data, format(data$DateTime_Case, "%Y") != "2023")
-  
   data <- format.table1(data)
   
   levels(data$ofi) <- c("OFI", "No OFI")
@@ -325,10 +324,8 @@ create.tableone <- function(data) {
       "ed_gcs_sum",
       "ed_sbp_value",
       "host_care_level",
-      "dt_ed_first_ct",
-      "dt_ed_emerg_proc",
       "res_survival",
-      "ed_emerg_proc"
+      "cohort"
     )
   
   render.continuous.new <- function(x) {
@@ -349,13 +346,13 @@ create.tableone <- function(data) {
   
   tableone <-
     table1(
-      ~ pt_age_yrs + pt_Gender + res_survival + host_care_level + ISS + ed_rr_value + ed_gcs_sum + ed_sbp_value + dt_ed_first_ct + dt_ed_emerg_proc + ed_emerg_proc
+      ~ pt_age_yrs + pt_Gender + res_survival + host_care_level + ISS + ed_rr_value + ed_gcs_sum + ed_sbp_value + cohort 
       | ofi,
       data = data[, vars2],
       render.missing = render.missing.new,
       render.continuous = render.continuous.new,
       render.categorical = render.categorical.new,
-      caption = "Demographic and Clinical Characteristics of patients screened for OFI.",
+      caption = "Demographic and clinical characteristics of the study sample.",
       #extra.col = list(`p-value` = pvalue),
       #extra.col.pos = 3
     )

### 2023-12-13 manuscript.Rmd
AI trauma intro, limitations, ethical

--- 2023-12-11 manuscript.Rmd
@@
 ---
-title: "Performance of prediction models for opportunities for improvment in trauma care among women compared to men and across patient cohorts"
+title: "Performance of prediction models for opportunities for improvment in trauma care across patient cohorts"
 author: "Anna Thoss"
@@
 ## Background
+Trauma is among the leading causes of mortality and morbidity worldwide.
+The selective review of patient's mortality and morbidity cases is used to improve quality of care through identification of opportunities for improvement (OFI).
+The aim of this study was to compare how machine learning models perform the prediction of OFI across different trauma cohorts.
 
 ## Methods
+We conducted a registry based study using all trauma patients at the Karolinska University Hospital from 2017 till 2021 that had undergone review for possible OFI.
+Previously trained machine learning models were assessed using statistical analysis to determine their prediction value across different patient cohorts.
 
@@
 # Introduction
-
-Trauma, defined as the clinical entity composed of the combination of physical injury and the body's associated responses [@gerdin_risk_2015] is a significant public health concern worldwide, accounting for a substantial portion of morbidity and mortality rates, being the cause of 4.3 million fatalities each year [@injuries2020].
+Trauma, defined as the clinical entity composed of the combination of physical injury and the body's associated responses [@gerdin_risk_2015] is a significant public health concern worldwide.
+It accounts for a substantial portion of morbidity and mortality rates, being the cause of 4.3 million fatalities each year [@injuries2020].
 It is the leading cause of death for people aged 18-39 [@Peng2023-gd].
@@
 
-#### Opportunities for improvement
-
+### Opportunities for improvement
 Audit filters, a system used to flag abnormal clinical parameters as possible opportunities for improvement (OFI), are applied to determine possible events in patient care that are associated with unfavourable outcomes.
@@
 
-#### Artificial intelligence
-
-In recent years, artificial intelligence (AI) has made great advances, making it a topic of interest in regard to healthcare applications.
+### Artificial intelligence
+In recent years, artificial intelligence (AI) has made great advances, making it an exciting opportunity for the development of  more sophisticated prediction models.
+This opens up a wide field of possible future healthcare applications.
 Research on AI implementation for clinical use has become commonplace and has already been tried in some clinical contexts, such as aiding in the diagnosis of pathologies through analysis of radiological images [@Cellina2022-qp] as well as epidemiological prediction of disease patterns, an example being during the COVID-19 pandemic [@Chen2020-ue].
@@
 ### AI in trauma care
+Following the general trend towards implementation of machine learning models within healthcare, there has been research into possible application within the wider field of trauma care.
+Previous methods to predict outcomes within trauma, such as Injury severity score (ISS) and Trauma and Injury severity score (TRISS) were developed using much smaller quantities of data than processing allows for these days[@Zhang2022], this has lead some to call their accuracy of prediction into question.[@Schetinin2018].
+There is an implicit heterogeneity to trauma patients, which lends itself well to machine learning modelling, having not only the ability to make predictions based on a greater dataset, but also circumvents problems as it can handle unstructured and nonlinear data and account for missing values.[@Zhang2022]
 
-<!-- Expand on this section to introduce AI/ML in trauma care more generally, and then focus on the application of ML to select patients for M&M -->
+Previous studies have found that delay of treatment along with errors in judgement are the leading causes of mortality in level 1 trauma centres.[@Teixeira2007;@Teixeira2009] Machine learning has been found to outperform clinical tools for prediction of mortality.[@Kareemi2021] Research on morbidity in trauma patients has been less conclusive, and fewer studies have been done on the topic, with most focusing on mortality.[@Zhang2022]
 
@@
 This is likely because these models were originally developed with a focus on predicting mortality rather than morbidity or instances of care failures within the realm of trauma care.
-
 Machine learning-based prediction models have been found to outperform conventional methods to predict opportunities for improvement [@Attergrim2023].
 
-#### Trauma care in different patient cohorts
+### Trauma care in different patient cohorts
+The array of injuries to the body that fall under trauma is wide.
+Following this, there is also a wide array of bodily responses to said injury.
+For instance, traumatic brain injury (TBI) will impact the body in different ways than an injury without TBI, since the body's response mechanism might be impaired by the neural injury in itself.
+This was shown BLAA BLAA BLAA.
+[Behöver KÄLLA] Along the same line, the mechanism of injury, such as blunt force trauma compared to severe penetrating trauma may also trigger different mechanisms and present with a different set of parameters clinical[BEHÖvER KÄLLA].
+Thus, it is important, that given the differences in cohorts falling under the wider term trauma patient, that the model used for predicting OFI does not discriminate or vary widely between the cohorts, so as not to leave out a subgroup of patients from optimising the care of said patients.
 
-<!-- Similar performance across different subgroups is also important from a fairness perspective, meaning that if the model performs worse among women, then we risk providing lower quality care to women, because we might miss important OFI -->
+- predicting outcomes is difficult, complexity. young group.
 
-Men make up a disproportionally large group compared to women when it comes to trauma.
-Each year, almost twice as many men die due to traumatic injury when compared to women, with the distribution of different types of injury varying by demographic and region [@WHO2014].
-Females have been shown to have a higher survival following trauma in higher income countries, whereas those in lower income countries show no difference compared to men.
-Differences in cell-mediated immune response following haemorrhage, differences in post-traumatic cytokine levels, and potential protective effects of oestrogen have been proposed to be the cause of the disparity between these groups [@Pendleton2022].
-It is important, that due to the differences between men and women, that the model for prediction has similar performance in these subgroups.
-Furthermore, it is important to establish if trained models perform equally well across different cohorts within the wider span of trauma, such as patients with traumatic brain injury (TBI) and without.
+# Aim
 
-## Aim
-
-To assess the difference in performance of prediction models for opportunities for improvement in trauma care among women compared to men and across clinical trauma cohorts.
+To assess the difference in performance of prediction models for opportunities for improvement in trauma care across different clinical trauma cohorts.
 
 # Methods
-
-## Methods
 
@@
 
+4.  Severe penetrating: At least one AIS \>2 injury in any of the following AIS body regions: neck, thorax, abdomen.
+
+5.  Other cohort: Patient does not fall into any of the above.
+
 Along with these cohorts, all machine learning models were assessed regarding differences between male and female sex in their prediction of OFI.
@@
 For each model, the performance was assessed in the predefined cohorts and compared to other models performance in regards to sensitivity and specificity.
+
+## Ethical considerations
+
+The nature of this study is non-invasive.
+However, handling of registry data constituts a risk for breach of patients integrity.
+The registry used in this study does contain personal identification.
+In order to minimize the risk of breach, patients have been semi-anonymized after collection of data and been assigned a new identification number upon entry into dataset.
+Due to the nature of the patient population, patients have not consented previous to entry into the registry, thus, there is an opt out principle applied here.
+This is common with registry data in Sweden, and opens up opportunities for studies in a wide range of fields.
+In turn, more scientific knowledge has great potential for benefit, both on the societal scale, as well as the individual.
+Thus, the risk for the individual is comparably small.
+Further, no individual patients data is presenteed in the study, mitigating risk of violating personal integrity.
+The study has been granted ethical permit, with the record number 2021-02541 and 2021-03531.
 
@@
 # Discussion
+### Key results
+- Summering av resultat + koppla till aim
+
+- Limitations:
+
+The potential of machine learning to aid clinicians in their efforts to improve trauma care quality and test the viability of its implementation in regards to application in different cohorts of trauma patients was at the forefront of this investigation.
+The result show that our models perform well in identifying OFI across the investigated cohorts.
+
+As this is a novel field, and research on similar applications of machine learning models to identify both mortality as well as morbidity is slim, especially in regard to OFI, it is hard to make a though discussion on how it compares to other findings in the field.
+However, a review from 2022 identified
+
+ <!-- Diskussionen kan läsas fristående och inleds därför
+med en kort sammanfattning av syfte eller
+problemställning, och viktigaste fynd
+
+Egna resultat analyseras och diskuteras i
+förhållande till studiens syfte
+
+Egna resultat analyseras och diskuteras i
+förhållande till stödjande resp motsägande aktuell
+forskning -->
+
+### Limitations
+
+<!-- Studiens (tekniska) styrkor och svagheter
+diskuteras (t ex val av metod/design och
+tillförlitlighet av metod och data) i eget stycke med
+särskild underrubrik -->
+
+The study had several limitations.
+Firstly, the nature of OFI is defined as a binary- variable.
+However, events leading up to a case being positive for OFI contains a multitude of variables, and can be determined by outcomes ranging from bad documentation to preventable death.
+Behind these are a diverse set of possible clinical events, which in turn can be determined by different predictors.
+
+Secondly, our learners are dependent on the current selection process for OFI, as this makes out the training data and thus makes out the groundwork for what is considered OFI in our dataset.
+As a consequence, the models are prone to the same limitations as the current system of audit filters.
+The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for demeriting OFI.
+This introduces an inherent bias.
+The audit filter system favour some types of error, while possibly missing others.
+At Karolinska University Hospital, the current system is more apt at identifying severe clinical outcomes, such as preventable death, or patient cases with more interventions or longer hospital stays, and flagging them as possible OFI.
+Higher ISS scores and "Days in hospital" have been found to be predictors of high importance when identifying OFI [Hussein].
+This in turn introduces possible selection bias, giving a sewed view of errors within the trauma care, and as such not a holistic view of the trauma care quality.
+
+Further, machine learning models are suboptimal when applied to cases with rare events, as they are limited to what they have previously been trained on.
+This means that not only are they limited when finding OFI in cases with novel predictors or events that lead with error, but the models are also prone to creating a self-reinforcing cycle, making them specialised on finding a set of OFI with accuracy.
+The current system can circumvent such by the individual review followed by a mortality and morbidity conference.
+If the model approach to predicting OFI is to be implemented in clinical practice, some other form, either individual review or other form of identification would be needed to supplement the machine learning model to assure prediction previously unknown basis for OFI.
+
+Finally, given that the aim of identifying OFI at mortality and morbidity conferences is the correction of errors within the system of trauma care, with time and implementation of corrective actions, the the scenarios and following this the parameters that predict OFI may change.
+
+### Interpretation
+
+### Generalisability
+
+<!-- Studiens praktiska implikationer eller betydelse
+avseende medicinvetenskapliga/samhälleliga
+koncept eller modeller diskuteras i eget stycke
+med särskild underrubrik. Beskriv också och
+problematisera aspekter av likabehandling inom
+det valda forskningsområdet. Om likabehandling
+inte anses vara relevant för studien ska detta
+motiveras -->
+
+"Machine learning has the potential for heavy
+bias, and the scrutiny of data sets, feature selection, and
+external validation are imperative for further
+implementation." - nr 18 review
+
+### Further research 
+<!-- Under rubriken Fortsatta studier ges konkreta
+förslag på nya hypoteser/frågeställningar och
+studiedesign som kan ge svar på dessa -->
+
+- Single center study > results illustrate situation in Stockholm. 
+- Difference in performance men vs women.
 

### 2023-12-13 manuscript.Rmd
Modify main results table and export as word file
can't remember

@@
 
-Karolinska trauma registry contains 11864 patients, of which 6310 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
+Karolinska trauma registry contains 11864 patients<!--Dubbelkolla antal!-->, of which 8309 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process <!--(n=2)-->.
+
 Through this, 1316 patients were selected for the morbidity conference, and 596 patients for the mortality conference.

### 2023-12-17 manuscript.Rmd
Discussion, minor changes in introduction, m&m

--- 2023-12-13 manuscript.Rmd
@@
 ## Background
+
 Trauma is among the leading causes of mortality and morbidity worldwide.
@@
 ## Methods
-We conducted a registry based study using all trauma patients at the Karolinska University Hospital from 2017 till 2021 that had undergone review for possible OFI.
-Previously trained machine learning models were assessed using statistical analysis to determine their prediction value across different patient cohorts.
 
@@
 # Introduction
+
 Trauma, defined as the clinical entity composed of the combination of physical injury and the body's associated responses [@gerdin_risk_2015] is a significant public health concern worldwide.
@@
 ### Opportunities for improvement
+
 Audit filters, a system used to flag abnormal clinical parameters as possible opportunities for improvement (OFI), are applied to determine possible events in patient care that are associated with unfavourable outcomes.
@@
 ### Artificial intelligence
+
 In recent years, artificial intelligence (AI) has made great advances, making it an exciting opportunity for the development of more sophisticated prediction models.
@@
 ### AI in trauma care
+
 Following the general trend towards implementation of machine learning models within healthcare, there has been research into possible application within the wider field of trauma care.
 Previous methods to predict outcomes within trauma, such as Injury severity score (ISS) and Trauma and Injury severity score (TRISS) were developed using much smaller quantities of data than processing allows for these days[@Zhang2022], this has lead some to call their accuracy of prediction into question.[@Schetinin2018].
-There is an implicit heterogeneity to trauma patients, which lends itself well to machine learning modelling, having not only the ability to make predictions based on a greater dataset, but also circumvents problems as it can handle unstructured and nonlinear data and account for missing values.[@Zhang2022]
+There is an implicit heterogeneity to trauma patients, which lends itself well to machine learning modelling, having not only the ability to make predictions based on a greater dataset, but also circumvents problems as it can handle unstructured and non-linear data and account for missing values.[@Zhang2022]
 
-Previous studies have found that delay of treatment along with errors in judgement are the leading causes of mortality in level 1 trauma centres.[@Teixeira2007;@Teixeira2009] Machine learning has been found to outperform clinical tools for prediction of mortality.[@Kareemi2021] Research on morbidity in trauma patients has been less conclusive, and fewer studies have been done on the topic, with most focusing on mortality.[@Zhang2022]
+Previous studies have found that delay of treatment along with errors in judgement are the leading causes of mortality in level 1 trauma centres.[@Teixeira2007; @Teixeira2009] Machine learning has been found to outperform clinical tools for prediction of mortality.[@Kareemi2021] Research on morbidity in trauma patients has been less conclusive, and fewer studies have been done on the topic, with most focusing on mortality.[@Zhang2022]
 
@@
 ### Trauma care in different patient cohorts
-The array of injuries to the body that fall under trauma is wide.
-Following this, there is also a wide array of bodily responses to said injury.
-For instance, traumatic brain injury (TBI) will impact the body in different ways than an injury without TBI, since the body's response mechanism might be impaired by the neural injury in itself.
-This was shown BLAA BLAA BLAA.
-[Behöver KÄLLA] Along the same line, the mechanism of injury, such as blunt force trauma compared to severe penetrating trauma may also trigger different mechanisms and present with a different set of parameters clinical[BEHÖvER KÄLLA].
-Thus, it is important, that given the differences in cohorts falling under the wider term trauma patient, that the model used for predicting OFI does not discriminate or vary widely between the cohorts, so as not to leave out a subgroup of patients from optimising the care of said patients.
 
-- predicting outcomes is difficult, complexity. young group.
+Trauma is a wide field with a heterogeneous patient population as well as encompassing an array of different injuries and responding mechanisms of the body to preserve homoeostasis.[@Dumovich2022] Trauma cases can be divided into cohorts in different ways, depending on mechanism of injury, damage to the nervous system or response of the body.
+Not only does the presentation of these patients wary, but management differers both due to severity of injury, as well as type of injury.
+[@Hornor2018] Thus, clinical judgement is an important component in treatment.
+As previous studies have not only found that judgement errors are common in OFI, but that the time frame between arrival at the emergency room and diagnostic tests such as CT or corrective procedures like surgery is longer in more critically ill patients [@Albaaj2023], the importances to not discriminate against a cohort in a machine learning model is vital in finding OFI for quality improvement across the board.
+If models are to be implemented in the clinical context in the future, it is important to establish that they perform well across different trauma cohorts.
 
@@
 
-#### Study setting and participants
+### Source of data
 
 We conducted a registry-based study, encompassing all trauma patients enlisted in both the Karolinska University Hospital trauma registry and the trauma care quality database spanning from 2012 to 2022.
+The trauma registry encompasses approximately 12,000 patients treated from 2012 to 2022.
+The trauma care quality database in turn is a subset of the trauma registry, including about 6000 patients selected for review in the years 2013-2022.
+
+### Participants
+
 The Karolinska University Hospital in Solna is equivalent to a level 1 trauma centre and manages approximately 1,500 acute trauma patients each year [@Ghorbani2018; @swetrau].
@@
 
-The trauma registry encompasses approximately 12,000 patients treated from 2012 to 2022.
-The trauma care quality database in turn is a subset of the trauma registry, including about 6000 patients selected for review in the years 2013-2022.
+All patients reviewed for possible OFI between the years 2013 and 2022 are included in the study, with the exception of patients under the age of 15 due to differing clinical pathways in this subgroup.
 
-<!-- Consider citing Hussein's paper in your definition of OFI: "The outcome was the presence of OFI, as decided by the morbidity and mortality conference.
+### Outcome
+
+The outcome in our study is OFI, as previously defined by Albaaj et.
+al: "The outcome was the presence of OFI, as decided by the morbidity and mortality conference.
 An OFI is any failure of care including, but not limited to, any potentially preventable or preventable death, delay in treatment, clinical judgment error, missed diagnosis and technical error as decided by the mortality and morbidity conference.
-The study outcome is binary with the levels “Yes - At least one OFI identified” and “No - No OFI identified”."
+The study outcome is binary with the levels "Yes - At least one OFI identified" and "No - No OFI identified"."[@Albaaj2023]
 
-https://sjtrem.biomedcentral.com/articles/10.1186/s13049-023-01157-y
-
--->
-
+An OFI is a consensus decision made during a conference and includes proposals for corrective actions.
 OFI are identified through meetings of the multidisciplinary review board during mortality and morbidity conferences.
@@
 
-![](assets/OFI-flow.png) Figure illustrating process of mortality and morbidity conference at Karolinska University Hospital.
+![](assets/OFI-flow.png) Figure illustrating process of mortality and morbidity conference at Karolinska University Hospital. <!--Note to self: Uppdatera!-->
 
-#### Eligibility criteria
+### Predictors
 
-All patients reviewed for possible OFI between the years 2013 and 2022 are included in the study, with the exception of patients under the age of 15 due to differing clinical pathways in this subgroup.
-
-#### Outcome
-
-The outcome of this study will be defined as "Yes, there is a significant difference in models prediction of OFI in different cohorts" or "No, there is no significant difference in models prediction of OFI in different cohorts." Data for this will be extracted from the trauma care quality database and run through a number of machine learning models, that have been trained previous to this study.
-
-#### Predictors
-
-All variables from the trauma registry are considered potential predictors in our analysis and have been incorporated into the models.
-<!-- Note that the models are already developed.
-So you should write about what have already been done,i.e.
-these variables are already incorporated into the
-models-->These variables include information spanning the pre-hospital, initial care, and subsequent in-hospital phases, which include initial vital signs, timing, and types of procedures and interventions, duration and level of care, injury details, mechanisms of injury, injury types, and standard demographic information.
-This array of predictors comprises both continuous and categorical variables, ultimately resulting in the final models utilizing a total of 45 predictors.
-
-All available data was used, with a sample size of 6310 patients.
-This is due to previously developed prediction models being based on all available data, as well as including all 45 predictors regardless of the learner used for the development of the model.
-
+All variables from the trauma registry are considered potential predictors in our analysis.
 The models have been developed and trained on the data set previous to this study [@Attergrim2023].
@@
 
-#### Cohorts
+Our variables include information spanning the pre-hospital, initial care, and subsequent in-hospital phases, which include initial vital signs, timing, and types of procedures and interventions, duration and level of care, injury details, mechanisms of injury, injury types, and standard demographic information.
+This array of predictors comprises both continuous and categorical variables, ultimately resulting in the final models utilizing a total of 45 predictors.
 
-The different models' performance were assessed in the following cohorts:
+### Sample size
+
+All available data was used, with a sample size of 8309 patients.
+This is due to previously developed prediction models being based on all available data, as well as including all 45 predictors regardless of the learner used for the development of the model.
+
+### Missing data
+
+<!-- Osäker vad jag ska skriva på den punkten -->
+
+### Cohorts
+
+The different models' performance were assessed in the following cohorts[@Hornor2018]:
 
@@
 
-4.  Severe penetrating: At least one AIS \>2 injury in any of the following AIS body regions: neck, thorax, abdomen.
+4.  Severe penetrating: At least one AIS \> 3 injury in any of the following AIS body regions: neck, thorax, abdomen.
 
@@
 
-Along with these cohorts, all machine learning models were assessed regarding differences between male and female sex in their prediction of OFI.
+<!-- skapa tabell för cohorter -->
 
-#### Statistical analysis
+### Statistical analysis
 
@@
 For each model, the performance was assessed in the predefined cohorts and compared to other models performance in regards to sensitivity and specificity.
+Area under the receiver operating curve (AUC) was used to measure discrimination.
 
-## Ethical considerations
+### Ethical considerations
 
 The nature of this study is non-invasive.
-However, handling of registry data constituts a risk for breach of patients integrity.
+However, handling of registry data constitutes a risk for breach of patients integrity.
 The registry used in this study does contain personal identification.
@@
 Thus, the risk for the individual is comparably small.
-Further, no individual patients data is presenteed in the study, mitigating risk of violating personal integrity.
+Further, no individual patients data is presented in the study, mitigating risk of violating personal integrity.
 The study has been granted ethical permit, with the record number 2021-02541 and 2021-03531.
 
-## Study design
+### Study design
 
@@
 
-## Setting
+### Setting
 
@@
 ```
-## Participants
+### Participants
 
@@
 
-Karolinska trauma registry contains 11864 patients, of which 6310 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
-Through this, 1316 patients were selected for the morbidity conference, and 596 patients for the mortality conference.
-Out of 596 deaths reviewed, 35 were found to be preventable (n=4) or possibly preventable (n=31).
-The remaining 561 patients were thus rendered non-preventable without OFI.
-Further, 1316 alive patients were included during a morbidity conference, of which 396 (7%) were identified as cases with OFI.
+Karolinska trauma registry contains 14022 patients, of which 8309 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
+Through this, 7571 patients were selected for the morbidity conference, and 727 patients for the mortality conference.
 
-Among the 6310 patients, the majority (n=4383) (69%) were male, and the mean age was 45 (SD 21). Overall, the mortality in the group was 599 (9%).
+Among the 8309 patients, the majority (n=5386) (69%) were male, and the mean age was 45 (SD 21).
+Overall, the mortality in the group was 727 (9%).
+In the group identified as OFI (n = 512), the mean age was somewhat higher than in the group without OFI (n=7797) (mean 49 vs 45 years).
+For OFI patients, they were most commonly treated at an intensive care unit (33%), whereas patients without OFI were most commonly treated at a general ward (38%).
+ISS was also higher for OFI patients (mean: 18, SD:11 vs mean: 12, SD:13).
+"Respiratory rate" was the most common missing variable at 20% (n=1644), followed by "ED GCS" at 10% (n=866).
 
-In the group identified as OFI (n = 431), the mean age was somewhat higher than in the group without OFI (n=5879) (mean 48 vs 45 years).
-For OFI patients, they were most commonly treated at an intensive care unit (35%), whereas patients without OFI were most commonly treated at a general ward (37%).
-ISS was also higher for OFI patients (mean: 19, SD:11 vs mean: 12, SD:13).
-In-hospital intubation occurred in 74 (17%) OFI patients, compared to non-OFI (n=460, 8%).
-They had a longer time till definitive treatment (median: 143 vs 99 minutes from hospital arrival).
-Definitive treatments differed, with OFI patients receiving more interventions, particularly radiological (7% vs 1%).
+Out of 727 deaths reviewed, 41 were found to be preventable or possibly preventable.
+The remaining 686 patients were thus rendered non-preventable without OFI.
+Further, out of the 7571 alive patients were included during a morbidity conference, 471 (6%) were identified as cases with OFI.
 
-The variables most frequently missing data were "Time to normal BE" (n=6107), "Pre-hospital intubation type" (N=5812), and "Emergency department intubation type" (n=5749).
-
-### Descriptive data
-
-### Outcome data
+The biggest cohort in the study was "other cohort" with 6707 (81%) of patients, followed by "blunt multisystem without TBI" at 9% (n=708).
+They also are the most common cohorts with OFI, making up 64% (n=329) and 23% (n=116) respectively.
+See table 1 for patient characteristics.
 
@@
 ```{r main.result, echo=FALSE}
-kable(cohort.performance.results, caption = "LÄGG TILL CAPTION HÄR")
+main.results.table <- flextable(cohort.performance.results) %>%
+  set_caption("Area under the curve (AUC) across clinical cohorts")
 ```
 
-<!-- KS: P-value was < 10-6 between the cohorts. Present that value where it suits you -->
+We compared the different cohorts with AUC as our performance metric.
+(See table 2) Isolated severe TBI had the highest AUC at 0.811 (0.807-0.814), followed by "Other cohort" (AUC 0.761 0.76-0.762).
+The model that performed the lowest in the cohort "Blunt multisystem with TBI" (AUC 0.696, 0.692-0.701) Figure 2 shows details of the receiver operating curve.
+The P-value between the different cohorts was 1 x 10 \^-6.
 
@@
 ```
-
-### Other analysis
 
@@
 # Discussion
+
 ### Key results
-- Summering av resultat + koppla till aim
 
-- Limitations:
+We set out to assess machine learning models viability as a tool to aid clinicians in their effort to improve trauma care quality, and specifically how they would perform across different trauma cohorts.
+We found that there is some variance in the models performance between the different cohorts, with "Blunt multisystem without TBI" performing the lowest, despite being the largest clearly defined trauma cohort in the study.
+The largest group "Other cohorts", which is made up of cases falling outside of the boundaries for the other formalized cohorts performs the best over all, despite possibly encompassing the largest heterogeneity.
 
-The potential of machine learning to aid clinicians in their efforts to improve trauma care quality and test the viability of its implementation in regards to application in different cohorts of trauma patients was at the forefront of this investigation.
-The result show that our models perform well in identifying OFI across the investigated cohorts.
+A previous study on the same dataset has found that audit filters, with an AUC of 0.624, are outperformed by our models when the dataset is not categorised by cohorts.[@Attergrim2023] It is unclear how audit filters perform in this context across cohorts as this was outside of the scope the study.
+Interestingly, the model performed better on the "Isolated severe TBI" cohort in terms of AUC then the best model in the aforementioned study (AUC 0.811 vs 0.789).
+A 2020 study found similarly good performance of machine learning models in TBI patients looking to predict poor outcomes.
+[@Matsuo2020] This may suggest that some models inclination to perform well in prediction of traumatic brain injury, whereas outcomes linked to other trauma cohorts may be harder to predict.
 
-As this is a novel field, and research on similar applications of machine learning models to identify both mortality as well as morbidity is slim, especially in regard to OFI, it is hard to make a though discussion on how it compares to other findings in the field.
-However, a review from 2022 identified
-
- <!-- Diskussionen kan läsas fristående och inleds därför
-med en kort sammanfattning av syfte eller
-problemställning, och viktigaste fynd
-
-Egna resultat analyseras och diskuteras i
-förhållande till studiens syfte
-
-Egna resultat analyseras och diskuteras i
-förhållande till stödjande resp motsägande aktuell
-forskning -->
+Overall the model perform well in the prediction of OFI across the cohorts.
+The application of machine learning models to outcomes within trauma is commonly more widely studied with more straight forward prediction of mortality, rather than the homogeneous reasons that make up the outcome of OFI.
+[@Zhang2022; @Gauss2023] Therefore, it is hard to make direct comparisons to other studies findings in this case, as they are generally either focused on a outcome far less complex, or not specifically tied to performance in outcomes in a particular cohort.
 
 ### Limitations
-
-<!-- Studiens (tekniska) styrkor och svagheter
-diskuteras (t ex val av metod/design och
-tillförlitlighet av metod och data) i eget stycke med
-särskild underrubrik -->
 
@@
 Behind these are a diverse set of possible clinical events, which in turn can be determined by different predictors.
+It has been suggested that data with higher resolution, such as vital signs series rather then single values could be a solution to this.
+However this comes with a drawback, as the current collection of data is based on the Upstein template (Uniform Reporting of Data following Major Trauma), and as such opens the door to a wider set of application in other contexts.
+Moving away from this has important implications for external validity of the model.
+[@Attergrim2023]
 
 Secondly, our learners are dependent on the current selection process for OFI, as this makes out the training data and thus makes out the groundwork for what is considered OFI in our dataset.
-As a consequence, the models are prone to the same limitations as the current system of audit filters.
+As a consequence, the models are prone to some of the same limitations as the current system of audit filters.
 The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for demeriting OFI.
@@
 At Karolinska University Hospital, the current system is more apt at identifying severe clinical outcomes, such as preventable death, or patient cases with more interventions or longer hospital stays, and flagging them as possible OFI.
-Higher ISS scores and "Days in hospital" have been found to be predictors of high importance when identifying OFI [Hussein].
-This in turn introduces possible selection bias, giving a sewed view of errors within the trauma care, and as such not a holistic view of the trauma care quality.
+Higher ISS scores and "Days in hospital" have been found to be predictors of high importance when identifying OFI [@Albaaj2023].
+This in turn introduces possible selection bias, giving a skewed view of errors within the trauma care, and as such not a holistic view of the trauma care quality.
 
@@
 Finally, given that the aim of identifying OFI at mortality and morbidity conferences is the correction of errors within the system of trauma care, with time and implementation of corrective actions, the the scenarios and following this the parameters that predict OFI may change.
+Inherent to machine learning models is the potential for heavy bias.[@Zhang2022] To mitigate this, a model should be subject to scrutiny of data sets and features selected, and external validation.
 
@@
 
-### Generalisability
+At this point it worth to note that perfect performance of the model was neither expected or the end goal.
+As previously stated, OFI is a heterogeneous outcome, and comparing a machine learning model to the current system of both quantitative screening and subsequent human reviews is in some ways unfair.
+The goal in developing a model to perform this analysis is its potential in aiding human efforts, to streamline the selection process for mortality and morbidity conferences.
+If implemented in a clinical context, it would be less resource intensive and could potentially be adjusted to new clinical patterns in the future that could be associated with OFI.
 
-<!-- Studiens praktiska implikationer eller betydelse
-avseende medicinvetenskapliga/samhälleliga
-koncept eller modeller diskuteras i eget stycke
-med särskild underrubrik. Beskriv också och
-problematisera aspekter av likabehandling inom
-det valda forskningsområdet. Om likabehandling
-inte anses vara relevant för studien ska detta
-motiveras -->
+In this sense, as it stands state of the art machine learning models have immense potential to aid in these types of selection processes.
+Time and costs are finite resources, and as such, there should always be the question of if they are adequately spent and if there is room for optimisation.
+This does not only apply to the selection process itself, but also to mortality and morbidity conferences.
+Therefore, efforts to optimise the selection of possible OFI to be discussed is imperative.
+With the current selection process resources are spent on discussing cases without OFI as the outcome.
+To limit the time spent on such cases would be a win.
 
-"Machine learning has the potential for heavy
-bias, and the scrutiny of data sets, feature selection, and
-external validation are imperative for further
-implementation." - nr 18 review
+A recent guideline stresses the need for more consideration of aspects such as usability, trust in algorithm output and human factor human factors to enable implementation of decision systems based on AI into clinical practice.
+[@Vasey2022] Key barriers to entry for these kinds of model into the medical community remain.
+
+In this study, we set out to investigate the viability of such models across cohorts.
+The findings suggest that not all patient types of patient cases perform as well in this context, and could need further improvement before implementation in selection of cases to be discussed at a conference.
+This is important from a fairness perspective, as the type of injury should not dictate the opportunity to find and correct errors in clinical practise.
 
 ### Further research
-<!-- Under rubriken Fortsatta studier ges konkreta
-förslag på nya hypoteser/frågeställningar och
-studiedesign som kan ge svar på dessa -->
 
-- Single center study > results illustrate situation in Stockholm. 
-- Difference in performance men vs women.
+Overall, further research is needed to increase performance of the models.
+Under optimal circumstances, any method to screen for possible OFI should minimise the set of false positives and also capture any true OFI, and as such minimise the clinical cost and time needed for trauma case quality improvement.
+
+It is important to highlight that these findings are based on single a single centre study, and as such only illustrates the situation in Stockholm.
+To establish a model that has validity in a wider context, further research would have to establish how the performance may differ in a different clinical context, and take a wider set of data and circumstances into account.
+
+Further, men make up a disproportionally large group compared to women when it comes to trauma.
+Each year, almost twice as many men die due to traumatic injury when compared to women, with the distribution of different types of injury varying by demographic and region [@WHO2014].
+Females have been shown to have a higher survival following trauma in higher income countries, whereas those in lower income countries show no difference compared to men.
+Differences in cell-mediated immune response following haemorrhage, differences in post-traumatic cytokine levels, and potential protective effects of oestrogen have been proposed to be the cause of the disparity between these groups [@Pendleton2022].
+Research on audit filters has found no correlation between audit filters and OFI.
+[@Albaaj2023] It is important, that due to the differences between men and women, that the model for prediction has similar performance in these subgroups, and does not discriminate.
+For future studies, it will be important to establish that any prediction model implemented in clinical practice performs equally well for both genders.
 
@@
 
-<!-- Give a short conclusion, corresponding to the aim -->
+State of the art machine learning models show promise for future implementation into a clinical context to predict OFI.
+Our findings show that although they overall perform well across cohorts, some disparity exists in the models accuracy across different cohorts.
 

### 2023-12-17 manuscript.Rmd
Edit and comment on manuscript

@@
 
-output.format <- knitr::opts_knit$get('rmarkdown.pandoc.to')
+output.format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
 
@@
 This is likely because these models were originally developed with a focus on predicting mortality rather than morbidity or instances of care failures within the realm of trauma care.
-Machine learning-based prediction models have been found to outperform conventional methods to predict opportunities for improvement [@Attergrim2023].
+Machine learning-based prediction models have been found to outperform conventional methods to predict opportunities for improvement, and the learner XGBoost showed particular promise [@Attergrim2023].
 
@@
 
-To assess the difference in performance of prediction models for opportunities for improvement in trauma care across different clinical trauma cohorts.
+To assess the difference in performance of a machine learning based prediction model for opportunities for improvement in trauma care across different clinical trauma cohorts.
 
@@
 The models have been developed and trained on the data set previous to this study [@Attergrim2023].
-Eight models are included, all being developed using the Tidymodels framework [@tidymodels].
+Eight models were included in the original study, all being developed using the Tidymodels framework [@tidymodels].
 These include logistic regression (LR), random forest (RF), decision tree (DT), support vector machine with a radial basis kernel (SVM), XGBoost, LightGBM, CatBoost, and k-nearest neighbor (k-NN).
+In this study, we focused on the model developed using XGBoost, as this model performed the best in the original paper.
 
@@
     kable_styling(latex_options = c("repeat_header", "condensed"), repeat_header_text = "(cont.)", font_size = 8) %>%
-    kableExtra::footnote(general = "\\\\textit{Definition of  abbreviations:} OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\\\\newline Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.", general_title = "",threeparttable =  TRUE, escape = FALSE)
+    kableExtra::footnote(general = "\\\\textit{Definition of  abbreviations:} OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\\\\newline Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.", general_title = "", threeparttable = TRUE, escape = FALSE)
 } else if (output.format == "docx") {
@@
 
-Karolinska trauma registry contains 14022 patients, of which 8309 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
+The Karolinska trauma registry includes 14022 patients, of which 8309 are reviewed for OFI, either through selection by individual review or application of audit filters, as well as an unknown process (n=2).
 Through this, 7571 patients were selected for the morbidity conference, and 727 patients for the mortality conference.
@@
 The biggest cohort in the study was "other cohort" with 6707 (81%) of patients, followed by "blunt multisystem without TBI" at 9% (n=708).
-They also are the most common cohorts with OFI, making up 64% (n=329) and 23% (n=116) respectively.
+These were also the cohorts with the most OFI, making up 64% (n=329) and 23% (n=116) respectively.
 See table 1 for patient characteristics.
@@
 
-We compared the different cohorts with AUC as our performance metric.
-(See table 2) Isolated severe TBI had the highest AUC at 0.811 (0.807-0.814), followed by "Other cohort" (AUC 0.761 0.76-0.762).
-The model that performed the lowest in the cohort "Blunt multisystem with TBI" (AUC 0.696, 0.692-0.701) Figure 2 shows details of the receiver operating curve.
-The P-value between the different cohorts was 1 x 10 \^-6.
+We compared the performance of the ML model in the different cohorts with AUC as our performance metric (See table 2).
+The ML model performed best in the cohort of patients with isolated severe TBI, in which it had an AUC of 0.811 (0.807-0.814), followed by "Other cohort" (AUC 0.761 0.76-0.762).
+The ML model performed worst in the cohort of patients with blunt multisystem trauma with TBI (AUC 0.696, 0.692-0.701).
+Figure 2 shows details of the receiver operating curves.
+The P-value between the different cohorts was 1 x 10 \^-6, meaning that there were significant differences in AUC between cohorts.
 
@@
 
-We set out to assess machine learning models viability as a tool to aid clinicians in their effort to improve trauma care quality, and specifically how they would perform across different trauma cohorts.
-We found that there is some variance in the models performance between the different cohorts, with "Blunt multisystem without TBI" performing the lowest, despite being the largest clearly defined trauma cohort in the study.
+We set out to assess the performance of a model based on the machine learner XGBoost's viability as a tool to aid clinicians in their effort to improve trauma care quality, and specifically how they would perform across different trauma cohorts.
+We found that there is some variation in the model's performance between the different cohorts, with "Blunt multisystem without TBI" performing the lowest, despite being the largest clearly defined trauma cohort in the study.
 The largest group "Other cohorts", which is made up of cases falling outside of the boundaries for the other formalized cohorts performs the best over all, despite possibly encompassing the largest heterogeneity.
 
-A previous study on the same dataset has found that audit filters, with an AUC of 0.624, are outperformed by our models when the dataset is not categorised by cohorts.[@Attergrim2023] It is unclear how audit filters perform in this context across cohorts as this was outside of the scope the study.
+A previous study on the same dataset has found that audit filters, with an AUC of 0.624, are outperformed by ML models when the dataset is not categorised by cohorts.[@Attergrim2023] It is unclear how audit filters perform in this context across cohorts as this was outside of the scope the study.
 Interestingly, the model performed better on the "Isolated severe TBI" cohort in terms of AUC then the best model in the aforementioned study (AUC 0.811 vs 0.789).
@@
 
-Secondly, our learners are dependent on the current selection process for OFI, as this makes out the training data and thus makes out the groundwork for what is considered OFI in our dataset.
+Secondly, our learner is dependent on the current selection process for OFI, as this makes out the training data and thus makes out the groundwork for what is considered OFI in our dataset.
 As a consequence, the models are prone to some of the same limitations as the current system of audit filters.
-The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for demeriting OFI.
+The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for demeriting<!-- Do you mean determining?--> OFI.
 This introduces an inherent bias.
@@
 
-In this study, we set out to investigate the viability of such models across cohorts.
-The findings suggest that not all patient types of patient cases perform as well in this context, and could need further improvement before implementation in selection of cases to be discussed at a conference.
+In this study, we set out to investigate the viability of a model based on the machine learner XGBoost across cohorts.
+The findings suggest that the model does not perform as well across all patient types, and could need further improvement before implementation in selection of cases to be discussed at a conference.
 This is important from a fairness perspective, as the type of injury should not dictate the opportunity to find and correct errors in clinical practise.

### 2023-12-18 manuscript.Rmd
something, i'm sure
add methodology and fix merge conflict

@@
 Our variables include information spanning the pre-hospital, initial care, and subsequent in-hospital phases, which include initial vital signs, timing, and types of procedures and interventions, duration and level of care, injury details, mechanisms of injury, injury types, and standard demographic information.
-This array of predictors comprises both continuous and categorical variables, ultimately resulting in the final models utilizing a total of 45 predictors.
+This array of predictors comprises both continuous and categorical variables, ultimately resulting in the final models utilizing a total of 17 predictors.
 
@@
 All available data was used, with a sample size of 8309 patients.
-This is due to previously developed prediction models being based on all available data, as well as including all 45 predictors regardless of the learner used for the development of the model.
+This is due to previously developed prediction models being based on all available data, as well as including all 17 predictors regardless of the learner used for the development of the model.
 
@@
 
-R was employed for all statistical analysis.
-For the analysis, a 95% confidence level and 5% significance level were employed.
-For each model, the performance was assessed in the predefined cohorts and compared to other models performance in regards to sensitivity and specificity.
-Area under the receiver operating curve (AUC) was used to measure discrimination.
+All statistical analyses were conducted using R.
+To extract probabilities for each patient case in the dataset, a 10-fold cross-validation strategy was employed, during which predictions for the holdout fold were recorded for each data point.
+We assessed performance within predefined cohorts and compared it across the different cohorts in terms of discrimination.
+This comparison was based on the area under the receiver operating characteristic curve (AUC).
+To estimate the 95% confidence interval for AUC performance, we employed bootstrapping on the extracted probabilities, repeating the process 1000 times.
+Additionally, an Analysis of Variance (ANOVA) test was conducted to compare differences between cohorts.
+A significance level of 5% was employed
 

### 2023-12-18 main.R

@@ -55,8 +55,6 @@ dataset <-  dataset %>%
   mutate(cohort = create_cohorts(.))
 
 # load results and merge into main dataframe.
-results <- readRDS("/opt/data/ml-ofi/random_predictions.rds")
-dataset <- merge(dataset, results$cat, by = "did", all.x = TRUE)
 results <- readRDS("/opt/data/ml-ofi/results/23-11-29-21-39_extract/results.rds")
 dataset <- dataset %>%
   left_join(results[[model.name]], by = "did") %>%
@@ -76,15 +74,7 @@ resample.calculate.auc <- function(cohort.data, bootstrap.iter) {
   roc_auc_vec(bootstrap.sample$ofi, bootstrap.sample$model_prediction, event_level = "second")
 }
 
-# Calculate AUC for the model as a whole
-binary_labels <- ifelse(dataset$ofi == "Yes", 1, 0)
-print(pROC::auc(binary_labels, dataset$repeat_1))
 main.results.path <- "results/main_results.rds"
-
-#### ex på hur man kan räkna AUC för blunt multisystem without Traumatic brain injury  
-
-bm.no.tbi <- dataset[dataset$cohort == "blunt multisystem without TBI",]
-
 
 if(!file.exists(main.results.path)) {
   
@@ -126,12 +116,12 @@ cohort.performance.results <- results %>%
                    cohort.ci[["Estimate"]], 
                    cohort.ci[["CI lower"]], 
                    cohort.ci[["CI upper"]]))
-    })) %>%
+  })) %>%
   mutate(N = map_int(cohort, function(cohort.name){
     return(cohort.datasets %>%
-      filter(cohort == cohort.name) %>%
-      unnest(data) %>%
-      nrow())
+             filter(cohort == cohort.name) %>%
+             unnest(data) %>%
+             nrow())
   })) %>%
   select(-auc) %>%
   rename(Cohort = cohort) %>%
@@ -186,4 +176,3 @@ cohort.roc <- ggroc(cohort.roc.objs) +
   annotate("text", x=0.41, y=0.485, label= "No discrimination", size = 4.5)
 
 ggsave("figures/roc.pdf", plot = cohort.roc, device = "pdf", height=10, width = 10)
-

### 2023-12-18 manuscript.Rmd

--- 2023-12-17 manuscript.Rmd
@@
 An OFI is any failure of care including, but not limited to, any potentially preventable or preventable death, delay in treatment, clinical judgment error, missed diagnosis and technical error as decided by the mortality and morbidity conference.
-The study outcome is binary with the levels "Yes - At least one OFI identified" and "No - No OFI identified"."[@Albaaj2023]
+The study outcome is binary with the levels"Yes - At least one OFI identified" and "No - No OFI identified"."[@Albaaj2023]
 

### 2024-01-08 Seminarieversion.Rmd
upload utkast slutversion

--- 2023-12-18 manuscript.Rmd
@@
 ---
-title: "Performance of prediction models for opportunities for improvment in trauma care across patient cohorts"
+title: "Performance of prediction models for opportunities for improvement in trauma care across patient cohorts"
 author: "Anna Thoss"
 output:
+  word_document: default
+  html_document: default
   pdf_document: default
-  html_document: default
-  word_document: default
 bibliography: bibliography.bib
@@
 
-```{r setup, include=FALSE, echo=FALSE}
-library(knitr)
-library(kableExtra)
-library(flextable)
-
-knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = NA)
-
-set_flextable_defaults(
-  font.size = 8,
-  theme_fun = theme_vanilla,
-  padding.top = 1,
-  padding.bottom = 1,
-  padding.left = 1,
-  padding.right = 1
-)
-
-output.format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
-
-source("./main.R")
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
-
-```{=html}
-<!-- This is an R Markdown document. Markdown is a simple formatting
-syntax for authoring HTML, PDF, and MS Word documents. For more
-details on using R Markdown see http://rmarkdown.rstudio.com -->
-```
-```{=html}
-<!-- This is a comment and will not be present in the compiled
-document. Try "knitting" this document by clicking the knit button up
-to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
-the console and you should only see headings -->
-```
-```{=html}
-<!-- Structure and most comments are from the Strengthening the
-reporting of observational studies in epidemiology (STROBE) statement,
-see https://www.strobe-statement.org/. For more explanations see the
-paper Strengthening the Reporting of Observational Studies in
-Epidemiology (STROBE):explanation and elaboration.  -->
-```
-```{=html}
-<!-- Indicate the study’s design with a commonly used term in the
-title, for example: "Associations between resuscitation interventions
-and opportunities for improvement in adult trauma patients: A cohort
-study" -->
-```
-# Abstract
-
-```{=html}
-<!-- Provide in the abstract an informative and balanced summary of
-what was done and what was found. Not more than 300 words. -->
-```
-## Background
-
-Trauma is among the leading causes of mortality and morbidity worldwide.
-The selective review of patient's mortality and morbidity cases is used to improve quality of care through identification of opportunities for improvement (OFI).
-The aim of this study was to compare how machine learning models perform the prediction of OFI across different trauma cohorts.
-
-## Methods
-
-## Results
-
-## Conclusion
 
@@
 
-![](assets/OFI-flow.png) Figure illustrating process of mortality and morbidity conference at Karolinska University Hospital. <!--Note to self: Uppdatera!-->
+Figure illustrating process of mortality and morbidity conference at Karolinska University Hospital.
 
@@
 This is due to previously developed prediction models being based on all available data, as well as including all 17 predictors regardless of the learner used for the development of the model.
-
-### Missing data
-
-<!-- Osäker vad jag ska skriva på den punkten -->
 
@@
 
-<!-- skapa tabell för cohorter -->
-
 ### Statistical analysis
@@
 Additionally, an Analysis of Variance (ANOVA) test was conducted to compare differences between cohorts.
-A significance level of 5% was employed
+A significance level of 5% was employed.
 
@@
 
-### Study design
-
-<!-- Present key elements of study design -->
-
-### Setting
-
-```{=html}
-<!-- Describe the setting, locations, and relevant dates, including
-periods of recruitment, exposure, follow-up, and data collection -->
-```
-### Participants
-
-```{=html}
-<!-- Cohort study: Give the eligibility criteria, and the sources and
-methods of selection of participants. Describe methods of
-follow-up. For matched studies, give matching criteria and number of
-exposed and unexposed -->
-```
-```{=html}
-<!-- Case-control study: Give the eligibility criteria, and the
-sources and + methods of case ascertainment and control
-selection. Give the rationale for the choice of cases and
-controls. For matched studies, give matching criteria and the number
-of controls per case -->
-```
-```{=html}
-<!-- Cross-sectional study: Give the eligibility criteria, and the
-sources and methods of selection of participants -->
-```
-## Variables and data sources/measurements
-
-```{=html}
-<!-- Clearly define all outcomes, exposures, predictors, potential
-confounders, and effect modifiers. Give diagnostic criteria, if
-applicable. For each variable of interest, give sources of data and
-details of methods of assessment (measurement). Describe comparability
-of assessment methods if there is more than one group -->
-```
-## Bias
-
-<!-- Describe any efforts to address potential sources of bias -->
-
-## Study size
-
-<!-- Explain how the study size was arrived at -->
-
-## Quantitative variables
-
-```{=html}
-<!-- Explain how quantitative variables were handled in the
-analyses. If applicable, describe which groupings were chosen and why
--->
-```
-## Statistical methods
-
-```{=html}
-<!--
-
-(a) Describe all statistical methods, including those used to control
-for confounding
-
-(b) Describe any methods used to examine subgroups and interactions
-
-(c) Explain how missing data were addressed
-
-(d)
-
-Cohort study: If applicable, explain how loss to follow-up was addressed
-
-Case-control study:If applicable, explain how matching of cases and
-controls was addressed
-
-Cross-sectional study: If applicable, describe analytical methods
-taking account of sampling strategy
-
-(e) Describe any sensitivity analyses
--->
-```
 # Results
@@
 ### Participants
-
-```{r tab1, echo=FALSE, fig.cap="Selected sample characteristics", fig.align='center' }
-if (output.format == "latex") {
-  tab1 <- t1kable(create.tableone(dataset), "latex", longtable = T, booktabs = T) %>%
-    kable_styling(latex_options = c("repeat_header", "condensed"), repeat_header_text = "(cont.)", font_size = 8) %>%
-    kableExtra::footnote(general = "\\\\textit{Definition of  abbreviations:} OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.\\\\newline Time to first CT and Time to definitive treatment: Measured in minutes from arrival at the hospital.", general_title = "", threeparttable = TRUE, escape = FALSE)
-} else if (output.format == "docx") {
-  tab1 <- t1flex(create.tableone(dataset)) %>%
-    add_footer_lines("OFI = Opportunity for Improvement; ED = Emergency Department; GCS = Glascow Coma Scale.") %>%
-    autofit()
-}
-
-tab1
-```
 
@@
 
-```{r main.result, echo=FALSE}
-main.results.table <- flextable(cohort.performance.results) %>%
-  set_caption("Area under the curve (AUC) across clinical cohorts")
-```
-
 We compared the performance of the ML model in the different cohorts with AUC as our performance metric (See table 2).
@@
 
-```{r roc.curves, out.width="80%",fig.cap="ROC Curves. LÄGG CAPTION HÄR", echo=FALSE, fig.align='center'}
-knitr::include_graphics("figures/roc.pdf", error = FALSE)
-```
-
-```{=html}
-<!--
-
-Participants
-------------
-
-(a) Report numbers of individuals at each stage of study—eg numbers
-potentially eligible, examined for eligibility, confirmed eligible,
-included in the study, completing follow-up, and analysed
-
-(b) Give reasons for non-participation at each stage
-
-(c) Consider use of a flow diagram
-
-Descriptive data
-----------------
-
-(a) Give characteristics of study participants (eg demographic,
-clinical, social) and information on exposures and potential
-confounders
-
-(b) Indicate number of participants with missing data for each
-variable of interest
-
-(c) Cohort study — Summarise follow-up time (eg, average and total
-amount)
-
-Outcome data
-------------
-
-Cohort study — Report numbers of outcome events or summary measures
-over time
-
-Case-control study — Report numbers in each exposure category, or
-summary measures of exposure
-
-Cross-sectional study — Report numbers of outcome events or summary
-measures
-
-Main results
-------------
-
-(a) Give unadjusted estimates and, if applicable, confounder-adjusted
-estimates and their precision (eg, 95% confidence interval). Make
-clear which confounders were adjusted for and why they were included
-
-(b) Report category boundaries when continuous variables were
-categorized
-
-(c) If relevant, consider translating estimates of relative risk into
-absolute risk for a meaningful time period
-
-Other analyses 
---------------
-
-Report other analyses done—eg analyses of subgroups and interactions,
-and sensitivity analyses 
--->
-```
-You can include code in this document like this:
-
-```{r main, echo=FALSE}
-source("main.R") ## This "imports" the main script file of your project and run any code in it
-```
-
-You can also embed plots:
-
-```{r plot, echo=FALSE}
-plot(pressure)
-```
-
-You can also mix text and code, so called inline code, like this: `r 2+5`.
-
 # Discussion
-
-### Key results
 
@@
 As a consequence, the models are prone to some of the same limitations as the current system of audit filters.
-The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for demeriting<!-- Do you mean determining?--> OFI.
+The review process is dependent both the individual review of specialized nurses, along with the current set of audit filters, which may not be the optimal parameters for determining OFI.
 This introduces an inherent bias.
@@
 
-```{=html}
-<!--
-
-Key results
------------
-Summarise key results with reference to study objectives
-
-Limitations
------------
-
-Discuss limitations of the study, taking into account sources of
-potential bias or imprecision.  Discuss both direction and magnitude
-of any potential bias
-
-Interpretation
---------------
-
-Give a cautious overall interpretation of results considering
-objectives, limitations, multiplicity of analyses, results from
-similar studies, and other relevant evidence
-
-Generalisability
-----------------
-
-Discuss the generalisability (external validity) of the study results
-
--->
-```
 # Conclusion
